---
title: "Simulated figures"
format: html
---

```{r setup}
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(patchwork)
library(ggseqlogo)
library(tidygraph)
library(ggraph)

# palette (single categorical hue, light mode) - see dataviz conventions
BLUE      <- "#2a78d6"
INK       <- "#0b0b0b"
INK_MUTED <- "#898781"
BASELINE  <- "#c3c2b7"
GRID_COL  <- "#e1e0d9"
SURFACE   <- "#fcfcfb"
```

## Simulated ChIP-seq peak

```{r chipseq-functions}
#' Simulate a ChIP-seq-like coverage track over a small genomic interval.
#'
#' Background reads are Poisson noise; the peak is an asymmetric
#' (skewed) Gaussian bump, smoothed with a moving average to mimic
#' fragment-extension / binning smoothing seen in genome browser tracks.
simulate_chipseq_track <- function(start = 1e6,
                                    length = 3000,      # bp span
                                    n_points = 1500,
                                    peak_center_frac = 0.45,
                                    peak_width = 180,   # bp
                                    peak_height = 48,
                                    background_mean = 3.0,
                                    skew = 0.35,
                                    seed = 7) {
  set.seed(seed)

  x <- seq(start, start + length, length.out = n_points)
  center <- start + peak_center_frac * length

  sigma_left  <- peak_width * (1 - skew)
  sigma_right <- peak_width * (1 + skew)
  sigma <- ifelse(x < center, sigma_left, sigma_right)
  peak_shape <- exp(-0.5 * ((x - center) / sigma)^2)

  signal_mean <- background_mean + peak_height * peak_shape
  raw_counts <- rpois(n_points, lambda = pmax(signal_mean, 0.05))

  window <- 15
  smoothed <- as.numeric(stats::filter(raw_counts, rep(1 / window, window), sides = 2))
  smoothed <- zoo::na.locf(smoothed, na.rm = FALSE)
  smoothed <- zoo::na.locf(smoothed, na.rm = FALSE, fromLast = TRUE)

  threshold <- background_mean + 0.35 * peak_height
  above <- smoothed > threshold
  if (any(above)) {
    call_start <- min(x[above])
    call_end   <- max(x[above])
  } else {
    call_start <- center - peak_width
    call_end   <- center + peak_width
  }
  summit_pos <- x[which.max(smoothed)]

  tibble(position = x, signal = smoothed) |>
    tidyr::drop_na() |>
    mutate(call_start = call_start, call_end = call_end, summit = summit_pos)
}

#' Plot a simulated ChIP-seq track: signal panel + called-peak bar beneath it.
plot_chipseq_track <- function(df, title = "Simulated ChIP-seq signal") {
  call_start <- df$call_start[1]
  call_end   <- df$call_end[1]
  summit_pos <- df$summit[1]
  summit_y   <- df$signal[which.min(abs(df$position - summit_pos))]

  p_track <- ggplot(df, aes(position, signal)) +
    geom_area(fill = BLUE, alpha = 0.18) +
    geom_line(color = BLUE, linewidth = 0.8) +
    annotate("point", x = summit_pos, y = summit_y, color = BLUE, size = 2,
             shape = 21, fill = BLUE, stroke = 0.6) +
    scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.03))) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(title = title, x = NULL, y = "Signal\n(reads per bin)") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(color = GRID_COL, linewidth = 0.3),
      axis.line.x        = element_blank(),
      axis.ticks         = element_blank(),
      axis.text          = element_text(color = INK_MUTED, size = 8),
      axis.title.y       = element_text(color = INK_MUTED, size = 9),
      plot.title         = element_text(color = INK, size = 12, hjust = 0),
      plot.background     = element_rect(fill = SURFACE, color = NA),
      panel.background    = element_rect(fill = SURFACE, color = NA),
      plot.margin         = margin(t = 5, r = 14, b = 5, l = 5)
    )

  p_call <- ggplot() +
    geom_rect(aes(xmin = call_start, xmax = call_end, ymin = 0.15, ymax = 0.85),
              fill = BLUE) +
    scale_x_continuous(labels = scales::comma,
                        limits = range(df$position), expand = expansion(mult = c(0, 0.03))) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Genomic position (bp)", y = "Peak\ncall") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid  = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks  = element_blank(),
      axis.text.x = element_text(color = INK_MUTED, size = 8),
      axis.title  = element_text(color = INK_MUTED, size = 9),
      axis.title.y = element_text(angle = 0, vjust = 0.5),
      plot.background  = element_rect(fill = SURFACE, color = NA),
      panel.background = element_rect(fill = SURFACE, color = NA),
      plot.margin      = margin(t = 5, r = 14, b = 5, l = 5)
    )

  p_track / p_call + plot_layout(heights = c(5, 1))
}
```

```{r chipseq-plot, fig.width=8, fig.height=3.6}
chipseq_df <- simulate_chipseq_track()
plot_chipseq_track(chipseq_df)
```

```{r}
ggsave(
  filename = "simulated_tf_chipseq_track.pdf",
  plot = plot_chipseq_track(chipseq_df),
  device = "pdf",
  width = 2.5, height = 1.2, units = "in"
)
```

## Simulated transcription factor motif logo

```{r motif-functions}
#' Simulate a position weight matrix (PWM) for a transcription factor motif.
#'
#' Conservation (the dominant base's probability at each position) tapers
#' from a conserved core to more degenerate flanks, like a real TF motif
#' (e.g. an E-box-like "CANNTG" core flanked by weaker positions).
simulate_pwm <- function(width = 10,
                          center_conservation = 0.85,
                          edge_conservation = 0.28,
                          seed = 42,
                          overrides = list()) {
  set.seed(seed)
  bases <- c("A", "C", "G", "T")
  mid <- (width + 1) / 2
  spread <- width / 4
  conservation <- edge_conservation +
    (center_conservation - edge_conservation) * exp(-0.5 * ((seq_len(width) - mid) / spread)^2)

  consensus_bases <- sample(bases, width, replace = TRUE)
  for (pos in names(overrides)) {
    consensus_bases[as.integer(pos)] <- overrides[[pos]]
  }

  pwm <- matrix(0, nrow = 4, ncol = width, dimnames = list(bases, NULL))
  for (i in seq_len(width)) {
    dom <- conservation[i]
    dom_idx <- which(bases == consensus_bases[i])
    other_idx <- setdiff(seq_along(bases), dom_idx)

    raw <- rgamma(3, shape = 2)
    split <- raw / sum(raw) * (1 - dom)

    probs <- numeric(4)
    probs[dom_idx] <- dom
    probs[other_idx] <- split
    pwm[, i] <- probs
  }
  pwm
}

#' Plot a true sequence logo (stretched letter glyphs, ggseqlogo) from a PWM.
#' If flank_ellipsis is TRUE, "..." is drawn just past both ends to signal
#' that the real sequence continues beyond the shown positions.
plot_seqlogo <- function(pwm, title = "Simulated TF motif", flank_ellipsis = FALSE) {
  consensus <- paste(rownames(pwm)[apply(pwm, 2, which.max)], collapse = "")
  width <- ncol(pwm)

  p <- ggseqlogo(pwm, method = "bits") +
    labs(title = NULL) +
    theme(
      axis.text.x        = element_blank(),
      axis.text.y        = element_blank(),
      axis.title.x       = element_blank(),
      axis.title.y       = element_blank(),
      axis.ticks.x       = element_blank(),
      axis.ticks.y       = element_blank(),
      axis.line          = element_blank(),
      plot.title         = element_text(color = INK, size = 12, hjust = 0),
      plot.background    = element_rect(fill = SURFACE, color = NA),
      panel.background   = element_rect(fill = SURFACE, color = NA)
    )

  if (flank_ellipsis) {
    p <- p +
      annotate("text", x = 0.1, y = 1, label = "...", size = 10,
               color = INK_MUTED, fontface = "bold") +
      annotate("text", x = width + 0.9, y = 1, label = "...", size = 10,
               color = INK_MUTED, fontface = "bold") +
      coord_cartesian(clip = "off") +
      theme(plot.margin = margin(t = 5, r = 20, b = 5, l = 20))
  }

  p
}
```

```{r motif-plot, fig.width=5, fig.height=3}
pwm <- simulate_pwm(width = 10)
plot_seqlogo(pwm)
```

```{r motif-save}
ggsave(
  filename = "simulated_tf_motif_logo.pdf",
  plot = plot_seqlogo(pwm),
  device = "pdf",
  width = 2.5, height = 1.2, units = "in"
)
```

A second, flatter motif: same seed (so it's the same underlying random draw as
the first), with position 6's dominant base swapped from T to A, and lower
conservation throughout so no position gets as tall.

```{r motif-plot-2, fig.width=5, fig.height=3}
pwm2 <- simulate_pwm(
  width = 10,
  center_conservation = 0.55,
  edge_conservation = 0.18,
  seed = 42,
  overrides = list(`6` = "A")
)
plot_seqlogo(pwm2)
```

```{r motif-save-2}
ggsave(
  filename = "simulated_tf_motif_logo_2.pdf",
  plot = plot_seqlogo(pwm2),
  device = "pdf",
  width = 2.5, height = 1.2, units = "in"
)
```

A third figure: the two logos above, combined side by side.

```{r motif-plot-combined, fig.width=8, fig.height=3}
plot_seqlogo(pwm) | plot_seqlogo(pwm2)
```

```{r motif-save-combined}
ggsave(
  filename = "simulated_tf_motif_logo_combined.pdf",
  plot = plot_seqlogo(pwm) | plot_seqlogo(pwm2),
  device = "pdf",
  width = 5, height = 1.2, units = "in"
)
```

A fourth, fully-conserved logo: every position has exactly one allele at
probability 1 (uniform 2 bits everywhere), rather than a mix of bases like
the first two -- a hard consensus sequence with no degeneracy.

```{r motif-consensus-functions}
#' Simulate a fully-conserved "hard" consensus PWM: probability 1 for one
#' base at every position, 0 for the other three -- uniform height (2 bits)
#' across the whole logo, no degeneracy at any position.
simulate_consensus_pwm <- function(width = 10, seed = 3, consensus = NULL) {
  set.seed(seed)
  bases <- c("A", "C", "G", "T")
  if (is.null(consensus)) consensus <- sample(bases, width, replace = TRUE)

  pwm <- matrix(0, nrow = 4, ncol = width, dimnames = list(bases, NULL))
  for (i in seq_len(width)) pwm[consensus[i], i] <- 1
  pwm
}
```

```{r motif-plot-consensus, fig.width=5, fig.height=3}
pwm_consensus <- simulate_consensus_pwm(width = 20)
plot_seqlogo(pwm_consensus, flank_ellipsis = FALSE)
```

```{r motif-save-consensus}
ggsave(
  filename = "simulated_tf_motif_logo_consensus.pdf",
  plot = plot_seqlogo(pwm_consensus, flank_ellipsis = FALSE),
  device = "pdf",
  width = 5.5, height = 1.2, units = "in"
)
```

## Simulated GWAS Manhattan plot

```{r manhattan-functions}
#' Simulate GWAS summary statistics across chromosomes.
#'
#' Per-SNP p-values are uniform under the null; a handful of association
#' peaks are injected as localized bumps in -log10(p) (mimicking an LD
#' block around a causal variant), similar in spirit to the ChIP-seq peak
#' simulation above.
simulate_gwas <- function(n_chr = 22,
                           snps_per_chr = 300,
                           n_peaks = 4,
                           peak_height = 6,
                           peak_width_frac = 0.02,
                           seed = 1) {
  set.seed(seed)
  chr_lengths <- round(3e8 * exp(-0.06 * seq_len(n_chr)))

  df <- purrr::map_dfr(seq_len(n_chr), function(chr) {
    len <- chr_lengths[chr]
    pos <- sort(sample(seq_len(len), snps_per_chr))
    tibble(chr = chr, pos = pos, neglog10p = -log10(runif(snps_per_chr)))
  })

  peak_chrs <- sample(seq_len(n_chr), n_peaks)
  for (chr_i in peak_chrs) {
    len <- chr_lengths[chr_i]
    peak_pos <- sample(seq_len(len), 1)
    width <- len * peak_width_frac
    idx <- which(df$chr == chr_i)
    d <- abs(df$pos[idx] - peak_pos)
    bump <- peak_height * exp(-0.5 * (d / width)^2) * runif(length(idx), 0.7, 1)
    df$neglog10p[idx] <- df$neglog10p[idx] + bump
  }

  chr_offsets <- setNames(c(0, cumsum(as.numeric(chr_lengths))[-n_chr]), seq_len(n_chr))
  df |>
    mutate(cum_pos = pos + chr_offsets[as.character(chr)],
           chr = factor(chr, levels = seq_len(n_chr)))
}

#' Plot a black-and-white Manhattan plot (points alternate ink/muted-ink by
#' chromosome; identity is carried by the x-axis chromosome labels, not color).
plot_manhattan <- function(df,
                            genome_wide_threshold = -log10(5e-8),
                            suggestive_threshold = -log10(1e-5),
                            title = NULL,
                            subtitle = "Genome-wide association") {
  chr_axis <- df |>
    group_by(chr) |>
    summarise(mid = (min(cum_pos) + max(cum_pos)) / 2, .groups = "drop")

  ggplot(df, aes(cum_pos, neglog10p, color = chr)) +
    geom_point(size = 0.7, alpha = 0.85, show.legend = FALSE) +
    # geom_hline(yintercept = genome_wide_threshold, linetype = "dashed",
    #            color = INK, linewidth = 0.4) +
    # geom_hline(yintercept = suggestive_threshold, linetype = "dotted",
    #            color = INK_MUTED, linewidth = 0.4) +
    scale_color_manual(values = rep(c(INK, INK_MUTED), length.out = nlevels(df$chr))) +
    scale_x_continuous(breaks = chr_axis$mid, labels = chr_axis$chr,
                        expand = expansion(mult = 0.01)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
    labs(title = title, subtitle = subtitle, x = "Chromosome", y = "-log10(pvalue)") +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_line(color = GRID_COL, linewidth = 0.3),
      axis.text.x   = element_text(size = 6, color = INK),
      axis.text.y   = element_blank(),
      axis.ticks.y  = element_blank(),
      axis.title.x  = element_text(color = INK, size = 9),
      axis.title.y  = element_blank(),
      plot.title    = element_text(color = INK, size = 12, hjust = 0),
      plot.subtitle = element_text(color = INK_MUTED, size = 10, hjust = 0),
      plot.background  = element_rect(fill = SURFACE, color = NA),
      panel.background = element_rect(fill = SURFACE, color = NA)
    )
}
```

```{r manhattan-plot, fig.width=9, fig.height=3.5}
gwas_df <- simulate_gwas()
plot_manhattan(gwas_df)
```


```{r}
ggsave(
  filename = "/beagle3/haky/users/temi/projects/TFXcan/misc/simulated_manhattan_plot.pdf",
  plot = plot_manhattan(gwas_df),
  device = "pdf",
  width = 5, height = 1.2, units = "in"
)
```

## Simulated multi-track epigenome locus view

```{r epigenome-functions}
#' Simulate a set of epigenomic tracks over one small locus: sharp promoter
#' and accessibility peaks, a broader active-enhancer mark, transcription
#' decaying across the gene body, a repressive domain and methylation that
#' anti-correlate with the active marks, and CTCF boundary peaks flanking
#' the locus -- the standard "browser track" view of a locus's epigenome.
simulate_epigenome_locus <- function(start = 1e6,
                                      length = 4000,
                                      n_points = 800,
                                      promoter_frac = 0.25,
                                      enhancer_frac = 0.65,
                                      seed = 11,
                                      smooth_window = 25) {
  set.seed(seed)
  x <- seq(start, start + length, length.out = n_points)
  bump <- function(x, center, width) exp(-0.5 * ((x - center) / width)^2)

  promoter <- start + promoter_frac * length
  enhancer <- start + enhancer_frac * length
  domain_center <- (promoter + enhancer) / 2
  domain_width  <- (enhancer - promoter)

  h3k4me3  <- 2 + 40 * bump(x, promoter, 60)
  h3k27ac  <- 3 + 15 * bump(x, promoter, 150) + 30 * bump(x, enhancer, 200)
  atac     <- 1 + 25 * bump(x, promoter, 40) + 20 * bump(x, enhancer, 50)
  cage     <- 1 + 45 * bump(x, promoter, 25)
  h3k27me3 <- 24 - 20 * bump(x, domain_center, domain_width)
  ctdf_left  <- start + 0.04 * length
  ctdf_right <- start + 0.94 * length
  ctcf     <- 1 + 35 * bump(x, ctdf_left, 50) + 32 * bump(x, ctdf_right, 50)

  smooth <- function(v, window = smooth_window) {
    s <- as.numeric(stats::filter(v, rep(1 / window, window), sides = 2))
    s <- zoo::na.locf(s, na.rm = FALSE)
    zoo::na.locf(s, na.rm = FALSE, fromLast = TRUE)
  }
  noisy_track <- function(mu) smooth(rpois(length(mu), lambda = pmax(mu, 0.05)))

  tibble(
    position = x,
    H3K4me3  = noisy_track(h3k4me3),
    H3K27ac  = noisy_track(h3k27ac),
    ATAC     = noisy_track(atac),
    CAGE     = noisy_track(cage),
    H3K27me3 = noisy_track(h3k27me3),
    CTCF     = noisy_track(ctcf)
  )
}

#' Stack the simulated tracks genome-browser-style: one row per mark, a
#' fixed categorical color per track (identity, not magnitude), no axes --
#' just the shapes, for a compact illustration.
plot_epigenome_tracks <- function(df) {
  track_colors <- c(H3K4me3 = BLUE, H3K27ac = "#1baf7a", ATAC = "#eda100",
                     CAGE = "#4a3aa7", H3K27me3 = "#e34948", CTCF = "#e87ba4")
  track_names <- names(track_colors)

  make_panel <- function(track) {
    ggplot(df, aes(x = position, y = .data[[track]])) +
      geom_area(fill = track_colors[[track]], alpha = 0.18) +
      geom_line(color = track_colors[[track]], linewidth = 0.7) +
      scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
      scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
      labs(x = NULL, y = track) +
      theme_minimal(base_size = 10) +
      theme(
        panel.grid   = element_blank(),
        axis.text    = element_blank(),
        axis.ticks   = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(color = INK_MUTED, size = 8, angle = 0,
                                     vjust = 0.5, hjust = 1),
        plot.background  = element_rect(fill = SURFACE, color = NA),
        panel.background = element_rect(fill = SURFACE, color = NA),
        plot.margin = margin(t = 2, r = 8, b = 2, l = 2)
      )
  }

  panels <- lapply(track_names, make_panel)
  patchwork::wrap_plots(panels, ncol = 1)
}
```

```{r epigenome-plot, fig.width=6, fig.height=6}
epi_df <- simulate_epigenome_locus()
plot_epigenome_tracks(epi_df)
```
```{r}
ggsave(
  filename = "/beagle3/haky/users/temi/projects/TFXcan/misc/simulated_epigenomes.pdf",
  plot = plot_epigenome_tracks(epi_df),
  device = "pdf",
  width = 2, height = 2.2, units = "in"
)
```

## Simulated network

```{r network-functions}
#' Simulate several small random (Erdos-Renyi) networks and scatter them
#' across a shared canvas. Each network gets its own force-directed layout,
#' then that layout is translated to a randomly placed (non-overlapping)
#' cluster center, so all networks can be drawn on one set of axes.
simulate_multi_network <- function(n_networks = 5,
                                    min_nodes = 4, max_nodes = 8,
                                    edge_prob = 0.45,
                                    canvas_size = 10,
                                    min_separation = 3.2,
                                    seed = 5) {
  set.seed(seed)

  centers <- tibble(cx = numeric(0), cy = numeric(0))
  while (nrow(centers) < n_networks) {
    cand <- c(runif(1, 0, canvas_size), runif(1, 0, canvas_size))
    ok <- nrow(centers) == 0 ||
      min(sqrt((centers$cx - cand[1])^2 + (centers$cy - cand[2])^2)) > min_separation
    if (ok) centers <- bind_rows(centers, tibble(cx = cand[1], cy = cand[2]))
  }

  nodes_list <- vector("list", n_networks)
  edges_list <- vector("list", n_networks)

  for (i in seq_len(n_networks)) {
    n_nodes <- sample(min_nodes:max_nodes, 1)
    g <- igraph::sample_gnp(n_nodes, p = edge_prob)
    xy <- igraph::layout_with_fr(g)

    nodes_list[[i]] <- tibble(
      network = i,
      id = seq_len(n_nodes),
      x = xy[, 1] * 0.6 + centers$cx[i],
      y = xy[, 2] * 0.6 + centers$cy[i]
    )

    em <- igraph::as_edgelist(g, names = FALSE)
    edges_list[[i]] <- if (nrow(em) > 0) {
      tibble(
        network = i,
        x    = nodes_list[[i]]$x[em[, 1]],
        y    = nodes_list[[i]]$y[em[, 1]],
        xend = nodes_list[[i]]$x[em[, 2]],
        yend = nodes_list[[i]]$y[em[, 2]]
      )
    } else {
      tibble(network = integer(0), x = double(0), y = double(0),
             xend = double(0), yend = double(0))
    }
  }

  list(
    nodes = bind_rows(nodes_list) |> mutate(network = factor(network)),
    edges = bind_rows(edges_list) |> mutate(network = factor(network, levels = levels(bind_rows(nodes_list)$network)))
  )
}

#' Plot the scattered networks: one fixed categorical hue per network so
#' each cluster's identity is unambiguous, muted edges, no axes or legend
#' (cluster membership is already visually obvious from color + position).
plot_multi_network <- function(net_data) {
  net_colors <- setNames(
    c(BLUE, "#1baf7a", "#eda100", "#4a3aa7", "#e34948", "#e87ba4", "#eb6834", "#008300"),
    NULL
  )[seq_len(nlevels(net_data$nodes$network))]
  names(net_colors) <- levels(net_data$nodes$network)

  ggplot() +
    geom_segment(data = net_data$edges,
                 aes(x = x, y = y, xend = xend, yend = yend, color = network),
                 linewidth = 0.3, alpha = 0.8, show.legend = FALSE) +
    geom_point(data = net_data$nodes, aes(x = x, y = y, color = network),
               size = 2, show.legend = FALSE) +
    scale_color_manual(values = net_colors) +
    coord_equal() +
    theme_void() +
    theme(
      plot.background  = element_rect(fill = SURFACE, color = NA),
      panel.background = element_rect(fill = SURFACE, color = NA)
    )
}
```

```{r network-plot, fig.width=6, fig.height=6}
multi_net <- simulate_multi_network()
plot_multi_network(multi_net)
```

```{r}
ggsave(
  filename = "/beagle3/haky/users/temi/projects/TFXcan/misc/simulated_networks.pdf",
  plot = plot_multi_network(multi_net),
  device = "pdf",
  width = 2, height = 2.2, units = "in"
)
```

## Simulated multilayer perceptron schematic

```{r mlp-functions}
#' Simulate a feed-forward MLP's node/edge layout: one column per layer,
#' nodes stacked and vertically centered within each column, and every
#' node in layer i connected to every node in layer i+1 (fully connected).
simulate_mlp <- function(layer_sizes = c(4, 6, 6, 3), layer_gap = 3, seed = 1) {
  set.seed(seed)
  n_layers <- length(layer_sizes)

  nodes <- purrr::imap_dfr(layer_sizes, function(n, layer) {
    tibble(layer = layer, node = seq_len(n),
           x = layer * layer_gap, y = seq_len(n) - (n + 1) / 2)
  })

  edges <- purrr::map_dfr(seq_len(n_layers - 1), function(i) {
    from <- nodes[nodes$layer == i, ]
    to   <- nodes[nodes$layer == i + 1, ]
    combos <- expand.grid(from_i = seq_len(nrow(from)), to_i = seq_len(nrow(to)))
    tibble(
      x0 = from$x[combos$from_i], y0 = from$y[combos$from_i],
      x1 = to$x[combos$to_i],     y1 = to$y[combos$to_i]
    )
  })

  list(nodes = nodes, edges = edges)
}

#' Plot the MLP schematic: one fixed categorical color per layer (column
#' position already conveys layer identity, so no legend), muted edges,
#' no axes -- a clean architecture diagram.
#' 
#' 
plot_mlp <- function(mlp_data) {
  n_layers <- length(unique(mlp_data$nodes$layer))
  layer_colors <- rep(c(BLUE, "#1baf7a", "#eda100", "#4a3aa7", "#e34948", "#e87ba4"),
                       length.out = n_layers)

  ggplot() +
    geom_segment(data = mlp_data$edges, aes(x = x0, y = y0, xend = x1, yend = y1), linewidth = 0.1, alpha = 0.7) +
    geom_point(data = mlp_data$nodes, aes(x = x, y = y, color = factor(layer)),
               size = 1, show.legend = FALSE, color = 'black', fill = "grey") +
    scale_color_manual(values = layer_colors) +
    coord_equal() +
    theme_void() +
    theme(
      plot.background  = element_rect(fill = SURFACE, color = NA),
      panel.background = element_rect(fill = SURFACE, color = NA)
    )
}
```

```{r mlp-plot, fig.width=7, fig.height=5}
mlp_data <- simulate_mlp(layer_sizes = c(4, 6, 6, 3))
plot_mlp(mlp_data)
```


```{r}
ggsave(
  filename = "/beagle3/haky/users/temi/projects/TFXcan/misc/simulated_mlp.pdf",
  plot = plot_mlp(mlp_data),
  device = "pdf",
  width = 1, height = 2, units = "in"
)
```