---
title: Connecting TFXcan's programs to gene regulatory networks
date: "Tues Feb 3 2026"
author: Temi
fig-format: png
fig-dpi: 800
---

# Introduction 

The aim of this project here is to test concordance of predicted binding with 

- gene regulatory networks 

- known TF binding-gene expression data e.g. co-expression

# Library loading and set up

```{r}
# cache.rebuid = TRUE, 
knitr::opts_chunk$set(fig.width = 3, fig.height = 3, fig.format = 'png', fig.dpi = 800, cache = TRUE, cache.rebuild = FALSE)
```
```{r initial definitions}
suppressMessages(library(tidyverse))
suppressMessages(library(glue))
```

```{r message=F, warning=F}
library(R.utils)
library(data.table)
library(qvalue)
library(ggpubr)
library(qqman)
library(rstatix)
library(ggplot2)
library(PRROC)
library(RColorBrewer)
library(rhdf5)
```

```{r}
qqunif.ggplot <- function(p, mlog10_p_thres=30, pch = NULL){
    require(ggplot2)
    ## thresholded by default at 1e-30
    p=na.omit(p)
    nn = length(p)
    xx =  -log10((1:nn)/(nn+1))
    
    p_thres = 10^{-mlog10_p_thres}
    if(sum(p < p_thres)){
      warning(paste("thresholding p to ", p_thres))
      p = pmax(p, p_thres)
    }

    dt <- data.table(pval = xx, dist = -sort(log10(p)))

    if(is.null(pch)){
        pch <- 19
    }

    pp <- ggplot2::ggplot(dt, aes(x = pval, y = dist)) +
        geom_point(pch = pch) +
        labs(x = expression(Expected~~-log[10](italic(p))),
          y = expression(Observed~~-log[10](italic(p)))) +
          geom_abline(aes(intercept = 0, slope = 1), col='gray', lty = 1)
        
    return(pp)
}

theme_simple <- theme_bw() + theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(),
    plot.title = element_text(family = 'Helvetica', colour = 'black', size = 9.33),
    text=element_text(size=9.33, colour = 'black', family = 'Helvetica'),
    axis.text = element_text(hjust = 1), 
    axis.text.x = element_text( hjust=0.5), 
    panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5)
)

qqcompare.ggplot <- function(p1, p2, names = c('observed', 'null'), mlog10_p_thres=30, pch = NULL){
    require(ggplot2)
    ## thresholded by default at 1e-30
    p1 = na.omit(p1)
    p2 = na.omit(p2)
    p_thres = 10^{-mlog10_p_thres}
    if(!is.null(p_thres)){pvec1 = pmax(p1, p_thres); pvec2 = pmax(p2, p_thres)}
    nn = length(pvec1)
    xx =  -log10((1:nn)/(nn+1))

    if(is.null(pch)){
        pch <- 19
    }

    pp <- ggplot2::ggplot() +
        geom_point(aes(x = xx, y = -sort(log10(pvec1)), col = names[1]), pch = pch) +
        geom_point(aes(x = xx, y = -sort(log10(pvec2)), col = names[2]), pch = pch) +
        labs(x = expression(Expected~~-log[10](italic(p))),
          y = expression(Observed~~-log[10](italic(p)))) +
          geom_abline(aes(intercept = 0, slope = 1), col='gray', lty = 1)
        
    return(pp)
}

ggplot_qqplot <- function(x, y, pch = NULL){
    qqdata <- qqplot(x = x, y = y, plot.it = F)
    if(is.null(pch)){
        pch <- 19
    }
    ggplot2::ggplot() +
        geom_point(aes(x = qqdata$x, y = qqdata$y), pch = pch)
}
```

I use some of the functions in these gists;

```{r warning=F, message=F}
devtools::source_gist('https://gist.github.com/TemiPete/250d9922b9516691f83bb1fd999a3ccc')
devtools::source_gist('https://gist.github.com/hakyim/38431b74c6c0bf90c12f')
devtools::source_gist('https://gist.github.com/hakyim/5d2251ea1a86009499e4ffdf47fe2735')
devtools::source_gist('https://gist.github.com/TemiPete/e7214eb308c4116e2f10ee96401be0c2')
devtools::source_gist('https://gist.github.com/TemiPete/b309a46e25ecec93127fb6756e68fb14')
devtools::source_gist('https://gist.github.com/TemiPete/9c487a30e2c1a27a1d46b68128c82d10')

devtools::source_gist("https://gist.github.com/TemiPete/d303781c7ddd9b6d8b4d0163804e80c2")
```


```{r}
tfxcan_directory <- '/beagle3/haky/users/temi/projects/TFXcan'
figures_directory <- "/beagle3/haky/users/temi/projects/experiments/gene_modules/figures"
```

```{r}
# knitr::knit_exit()
```


# Main question: can we link TFXcan TF programs to gene regulatory networks?


## What gene regulatory network? GRAND gene regulatory networks

https://grand.networkmedicine.org/

I downloaded 

```{r}
grn_directory <- '/scratch/beagle3/temi/GRNs'
```

Focus on protein-coding genes and their TSS
```{r}
tss_metadata <- data.table::fread("/beagle3/haky/users/temi/projects/dgtex/files/canonical_tss.matched_genes.tsv")
# remove duplicates
dups <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus) %>%
    dplyr::group_by(locus) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt)) %>% 
    dplyr::filter(cnt > 1)

dt.tss_metadata <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus) %>%
    dplyr::filter(!locus %in% dups$locus) %>% 
    dplyr::distinct()
```

```{r}
dt_LCL_grn <- data.table::fread(file.path(grn_directory, 'PANDA_LCL_Yoruba.csv')) 
dim(dt_LCL_grn)
dt_LCL_grn <- dt_LCL_grn %>% dplyr::select(any_of(c('ID', dt.tss_metadata$ensembl_gene_id)))
dim(dt_LCL_grn); dt_LCL_grn[1:5, 1:10]
```

```{r}
length(unique(colnames(dt_LCL_grn))); length(colnames(dt_LCL_grn))
```

```{r}
mat_LCL_grn <- tibble::column_to_rownames(dt_LCL_grn, 'ID')
mat_LCL_grn[1:5, 1:5]; dim(mat_LCL_grn)
```

```{r}
# dt_LCL_grn <- dt_LCL_grn %>%
#     tidyr::pivot_longer(cols = !ID, names_to = 'gene', values_to = 'connectivity')
# dim(dt_LCL_grn); dt_LCL_grn[1:5, ]
```

```{r}
# dt_LCL_grn %>%
#     dplyr::group_by(ID) %>%
#     dplyr::summarize(sumC = sum(connectivity > 0)) %>%
#     dplyr::arrange(desc(sumC))
```

```{r}
# dt_LCL_grn %>%
#     dplyr::group_by(ID) %>%
#     dplyr::filter(connectivity > 0) %>%
#     dplyr::summarize(meanC = mean(connectivity)) %>%
#     dplyr::arrange(desc(meanC))
```

#### What is the average connection Z-score across TFs
```{r}
avg.tf_connection <- base::rowMeans(mat_LCL_grn)
summary(avg.tf_connection)
```
```{r}
min.tf_connection <- apply(mat_LCL_grn, 1, min)
summary(min.tf_connection)
```
```{r}
max.tf_connection <- apply(mat_LCL_grn, 1, max)
summary(max.tf_connection)
```

```{r}
median.tf_connection <- apply(mat_LCL_grn, 1, median)
summary(median.tf_connection)
```

```{r}
sd.tf_connection <- apply(mat_LCL_grn, 1, sd)
summary(sd.tf_connection)
```


# Does binding of a transcription factor at the tss of a gene correlate with the expression of that gene?

## Download gene expression for Geuvadis indiviuals: https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-GEUV-1#processed-data
## https://ftp.ebi.ac.uk/biostudies/fire/E-GEUV-/001/E-GEUV-1/Files/E-GEUV-1/analysis_results/GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz

Data:

1. Gene expression in blood for Geuvadis individuals

2. Predicted TF binding at TSS of protein-coding genes for the same individuals above

## Predicted TF binding at TSS of protein-coding genes for Geuvadis

First, I need to predict TF binding at the TSS

- 

In the steps below, I prepare the data
```{r, eval = F}
# I need to know the loci I want to predict for. I will read for one individual and use that information.
tre <- h5ls("/beagle3/haky/data/Geuvadis_TSS_enformer_prediction/HG00116.h5")
H5close()

# Match tss names
tss_metadata <- data.table::fread("/beagle3/haky/users/temi/projects/dgtex/files/canonical_tss.matched_genes.tsv")

# 
tss_predictions <- gsub('_predictions', '', tre$name)

# 
valid_tss <- intersect(tss_predictions, tss_metadata$locus)

# loci
dt_loci <- data.table::data.table(valid_tss) %>% setnames(c('tss'))
dt_loci |> head()

dt_loci %>%
    data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.tss.tsv"), row.names = F, col.names = F, quote = F, sep = '\t')

# This curates the individuals
gexpr <- data.table::fread('/beagle3/haky/data/Geuvadis/expression/GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz')

# individuals
dt_ids <- colnames(gexpr)[-c(1:4)] %>%
    data.table::data.table() %>% setnames(c('individual'))
head(dt_ids)

dt_ids %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.individuals.tsv"), row.names = F, col.names = F, quote = F, sep = '\t')
```


- This predicts TF binding
```{r}
llist <- file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.tss.tsv")
ilist <- file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.individuals.tsv")
weights <- file.path(tfxcan_directory, "data/chippeaks/weights/ENPACT_CHIPPEAKS_2026-01-02.compiled_weights.lambda.1se.txt.gz")
#file.path(tfxcan_directory, 'data/enpact/weights/ENPACT_734_2025-04-24.compiled_weights.with_intercept.lambda.1se.txt.gz')
input_directory <- '/beagle3/haky/data/Geuvadis_TSS_enformer_prediction'
output_directory <- '/scratch/beagle3/temi/enpact_scores/geuvadis_tss_chippeaks_models'

dir.create(output_directory, recursive = TRUE)

glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/enpact_predict_hdf5.sbatch {llist} {weights} {ilist} {input_directory} {output_directory}")
```


- This collects into one rds.gz file

```{r}
wgts <- data.table::fread(file.path(tfxcan_directory, "data/chippeaks/weights/ENPACT_CHIPPEAKS_2026-01-02.compiled_weights.lambda.1se.txt.gz"))

dim(wgts)

gg <- strsplit(colnames(wgts), '_')
tfs <- sapply(gg, getElement, 1)
tissues <- sapply(gg, function(x){x[2]})

unique(tfs) |> length(); unique(tissues) |> length()
```

```{r}
ilist <- file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.individuals.tsv")
efile <- file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.lcl.gene_expression.tsv.gz")
predictions_directory <- '/scratch/midway3/temi/enpact_scores/geuvadis_tss_chippeaks_models'
output_file <- '/scratch/midway3/temi/enpact_scores/geuvadis_tss.enpact_scores.chippeak_models.rds.gz'

glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/collect_predictions_into_rds.sbatch {ilist} {efile} {predictions_directory} {output_file}")
```


## Gene expression in blood for Geuvadis individuals 

Compile and correlate with observed gene expression

```{r, eval= F}
gexpr <- data.table::fread('/beagle3/haky/data/Geuvadis/expression/GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz')

mat_gexpr <- gexpr %>%
    tidyr::separate_wider_delim(Gene_Symbol, delim = '.', names = c('gene_name', 'id')) %>% dplyr::select(-c(TargetID, id, Chr, Coord)) %>%
    dplyr::inner_join(dt.tss_metadata, by = c('gene_name' = 'ensembl_gene_id')) %>%
    dplyr::relocate(gene_name, locus) %>% dplyr::select(-gene_name) %>%
    tibble::column_to_rownames('locus') %>% as.matrix()

mat_gexpr[1:5, 1:5]; dim(mat_gexpr)

mat_gexpr %>% as.data.table(keep.rownames = 'gene') %>%
    data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.lcl.gene_expression.tsv.gz"), sep = '\t', col.names = T, row.names = F, quote = F, compress = 'gzip')
```

I need to know the common individuals. 
```{r}
inds.binding <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.individuals.tsv"), header =F)$V1

inds.expression <- colnames(gexpr)[-c(1:4)]

length(inds.expression); length(inds.binding)
```

```{r}
identical(sort(inds.expression), sort(inds.binding))
```

Correlation with gene expression
```{r}

#'/scratch/midway3/temi/enpact_scores/expression_vs_binding.enpact_scores.geuvadis.chippeak_models.data.rds.gz'

#'/scratch/midway3/temi/enpact_scores/geuvadis_tss.enpact_scores.chippeak_models.rds.gz'
efile <- file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.lcl.gene_expression.tsv.gz")
binding_matrix <- '/scratch/midway3/temi/enpact_scores/geuvadis_tss.enpact_scores.chippeak_models.rds.gz'
output_file_basename <- '/scratch/midway3/temi/enpact_scores/expression_vs_binding.enpact_scores.geuvadis.chippeak_models'

glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/correlate_binding_expression.sbatch {binding_matrix} {efile} {output_file_basename}")
```

### Analysis of the correlation with gene expression
```{r, cache.lazy = FALSE}
regr.expr_binding <- data.table::fread('/scratch/midway3/temi/enpact_scores/expression_vs_binding.enpact_scores.geuvadis.chippeak_models.regression.tsv.gz') %>%
    dplyr::filter(term == 'MX[i, ]') %>%
    dplyr::rename(estimate = Estimate, se = `Std. Error`, tvalue = `t value`, pvalue = `Pr(>|t|)`) %>%
    dplyr::mutate(zscore = estimate/se) %>%
    tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F)
dim(regr.expr_binding); regr.expr_binding[1:5, ]
```

```{r}
regr.expr_binding$tf |> unique() |> length()
```

```{r, cache.lazy = FALSE}
corr.expr_binding <- data.table::fread('/scratch/midway3/temi/enpact_scores/expression_vs_binding.enpact_scores.geuvadis.chippeak_models.correlation.tsv.gz') %>%
    dplyr::rename(tf_tissue = tf) %>%
    tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F)
dim(corr.expr_binding); corr.expr_binding[1:5, ]
```

```{r}
corr.expr_binding$tf |> unique() |> length()
```

### Are the p values of the correlation and regression equivalent?

```{r, eval = F}
ppn_concordance <- ggplot_qqplot(x = regr.expr_binding$pvalue, y = corr.expr_binding$spearman_pvalue, pch = '.') +
geom_abline(intercept = 0, slope = 1, col = 'red') +
theme_simple +
labs(y = 'Correlation', x = 'Regression', subtitle = 'QQ plot of p values')

ggsave(file.path(figures_directory, 'regression_vs_correlation.QQplot.pvalues.png'), ppn_concordance, width = 3000, height = 3000, dpi = 800, units = 'px')
```


## Is TF binding correlated with gene expression?

```{r}
pphist <- ggplot() +
    geom_histogram(aes(x = corr.expr_binding$spearman_corr), col = 'black') +
    labs(x = 'Correlation Coefficient', y = 'Number of TF binding-Target gene pairs') + theme_simple

ggsave(file.path(figures_directory, 'binding_expression.correlation.coefficients.histogram.png'), pphist, width = 3000, height = 3000, dpi = 800, units = 'px')
```

```{r, eval = F}
pnqq <- qqunif.ggplot(corr.expr_binding$spearman_pvalue, pch = '.') +
    theme_simple + labs(x = expression("Expected -log"[10]*"P"), y = expression("Observed -log"[10]*"P"))

ggsave(file.path(figures_directory, 'binding_expression.correlation.pvalues.QQplot.png'), pnqq, width = 3000, height = 3000, dpi = 800, units = 'px')
```

## To compare with the GRAND GRN, I will use the regression Z-scores. 

First, check that the estimates are not weird.

```{r}
summary(regr.expr_binding$estimate)
```
```{r, eval = F}
pp_coefs <- ggplot() +
    geom_histogram(aes(regr.expr_binding$estimate), col = 'black') +
    theme_simple +
    labs(y = 'Number of TF-tissue-gene trios', x = 'Beta')

ggsave(file.path(figures_directory, 'histogram.tf_tissue_genes.regression_coefficients.png'), pp_coefs, width = 3000, height = 3000, dpi = 800, units = 'px')

pp_coefs
```

```{r, eval = F}
pp_zscore <- ggplot() +
    geom_histogram(aes(regr.expr_binding$zscore), col = 'black') +
    theme_simple +
    labs(y = 'Number of TF-tissue-gene trios', x = 'Z-score')

ggsave(file.path(figures_directory, 'histogram.tf_tissue_genes.regression_zscores.png'), pp_zscore, width = 3000, height = 3000, dpi = 800, units = 'px')

pp_zscore
```



What are the most significant TF-tissue-gene trios?
```{r}
signif.binding_expr <- regr.expr_binding %>% 
    dplyr::group_by(tf_tissue) %>%
    dplyr::mutate(fdr_sig = qvalue::qvalue(pvalue, fdr.level = 0.05)$significant) %>% 
    dplyr::filter(fdr_sig == TRUE)
```

```{r}
head(signif.binding_expr); dim(signif.binding_expr)
```

```{r}
dt.count_signif <- signif.binding_expr %>%
    summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt))

dt.count_signif
```

```{r, eval = F}
dd <- dt.count_signif %>%
    head(n = 20) %>%
    tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F) %>%
    dplyr::arrange(desc(cnt)) %>%
    dplyr::mutate(TF_tissue = factor(tf_tissue, level = tf_tissue))

pp_number_significant <- dd %>%
    ggplot(aes(x = TF_tissue, y = cnt, fill = tissue)) +
    theme_simple +
    geom_col(col = 'black', width = 0.8, position = 'dodge') +
    theme(legend.position = "none") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(subtitle = '', x = 'TF-Tissue', y = 'Number of Significantly Correlated\nBinding-Expression Pairs')


ggsave(file.path(figures_directory, 'barplot.tf_tissue_genes.top_number_significant.png'), pp_number_significant, width = 5000, height = 3500, dpi = 800, units = 'px')

pp_number_significant
```

```{r}
ggplot() +
    geom_boxplot(aes(y = dt.count_signif$cnt)) +
    theme_simple +
    labs(y = 'Number of Significantly Correlated\nBinding-Expression Pairs')
```


Are the genes all common?

```{r, eval = F}
grps.signif.binding_expr <- split(signif.binding_expr, signif.binding_expr$tf_tissue)
```
```{r, eval = F}
genes_per_tfTissue <-  purrr::map(grps.signif.binding_expr, function(xx){
    xx$gene
})
names(genes_per_tfTissue) <- names(grps.signif.binding_expr)

common_genes <- purrr::reduce(genes_per_tfTissue, base::intersect)
length(common_genes)
```

::: {.callout}
These are distinct TSS across all TF-tissue pairs.
:::


### Correlation with Gene regulatory network

```{r, eval = F}
dt.ort_regr <- purrr::map(unique(regr.expr_binding$tf_tissue), function(ii){
    tryCatch({
        dta <- regr.expr_binding %>% dplyr::filter(tf_tissue == ii)
        tf <- dta$tf |> unique()
        dt_tfb <- dta %>%
            dplyr::inner_join(dt.tss_metadata, by = c('gene' = 'locus'))
        
        cm_genes <- intersect(colnames(dt_LCL_grn), dt_tfb$ensembl_gene_id)

        tf_expr_aa <- dt_tfb %>%
            dplyr::filter(ensembl_gene_id %in% cm_genes) %>% dplyr::select(gene = ensembl_gene_id, zscore)
    
        network_aa <- dt_LCL_grn %>% 
            dplyr::select(all_of(c('ID', cm_genes))) %>% 
            dplyr::filter(ID == tf) %>% 
            data.table::transpose(., keep.names = "ID", make.names = "ID") %>% 
            setnames(c('gene', 'connectivity'))
        mgd <- dplyr::inner_join(tf_expr_aa, network_aa, by = 'gene')
        return(mgd)
        
    }, error = function(err){
        return(NULL)
    })
    
}, .progress = TRUE)

names(dt.ort_regr) <- unique(regr.expr_binding$tf_tissue)
dt.ort_regr <- Filter(Negate(is.null), dt.ort_regr)


dt.ort_regr %>%
    dplyr::bind_rows(., .id = 'tf_tissue') %>%
    data.table::fwrite(., file.path(tfxcan_directory, 'experiments/gene_modules/files/tf_binding_expression.connectivity.data.chippeak_models.tsv.gz'), sep = '\t', compress = 'gzip', quote = F, col.names = T, row.names = F)
```

### Correlation for TF-Gene connectivity with Z-scores > 0 (positive)
```{r}
dt.ort_regr <- data.table::fread(file.path(tfxcan_directory, 'experiments/gene_modules/files/tf_binding_expression.connectivity.data.chippeak_models.tsv.gz')) 

list.ort_regr <- dt.ort_regr %>% 
    base::split(., .$tf_tissue)
```

```{r}
result.grn_concordance <- purrr::map(names(list.ort_regr), function(ii){
    gt <- list.ort_regr[[ii]]
    ct <- with(gt, cor.test(abs(zscore), connectivity))
    return(data.table(tf_tissue = ii, pvalue = ct$p.value, corr = ct$estimate))
}, .progress = TRUE) %>% dplyr::bind_rows() %>% tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F)

result.grn_concordance[1:5, ]; dim(result.grn_concordance)
```


```{r}
pp_qqplot <- qqunif.ggplot(result.grn_concordance$pvalue) +
    theme_simple + labs(x = expression("Expected -log"[10]*"P"), y = expression("Observed -log"[10]*"P"), subtitle = 'p values')

ggsave(file.path(figures_directory, 'GRN_concordance.pvalues.QQplot.absZscore.png'), pp_qqplot, width = 2000, height = 2000, dpi = 800, units = 'px')
pp_qqplot
```

```{r}
summary(result.grn_concordance$corr)
```

The correlation is not that high. 

```{r}
result.grn_concordance.aggregated <- result.grn_concordance %>% dplyr::group_by(tf) %>% dplyr::summarize(acatP = ACAT::ACAT(pvalue))

result.grn_concordance.aggregated[1:5, ]
```


```{r, eval = T}
pp_qqplot <- qqunif.ggplot(result.grn_concordance.aggregated$acatP) +
    theme_simple + labs(x = expression("Expected -log"[10]*"P"), y = expression("Observed -log"[10]*"P"), subtitle = 'ACAT aggregated p values')

ggsave(file.path(figures_directory, 'GRN_concordance.acatP.QQplot.absZscore.png'), pp_qqplot, width = 2000, height = 2000, dpi = 800, units = 'px')
pp_qqplot
```

How many are significant after multiple testing correction, FDR < 0.05?

```{r}
result.grn_concordance.signif <- result.grn_concordance %>%
    dplyr::mutate(fdr_signif = qvalue::qvalue(pvalue, fdr.level = 0.05)$significant)
```

```{r}
result.grn_concordance.signif$fdr_signif %>% table() %>% prop.table()
```

How many are truly correlated?
```{r}
1 - qvalue::qvalue(result.grn_concordance$pvalue, fdr.level = 0.05)$pi0
```




### Examples

```{r}
top12 <- result.grn_concordance.signif %>%
    dplyr::filter(fdr_signif == TRUE) %>%
    dplyr::arrange(desc(abs(corr))) %>%
    dplyr::slice_head(n = 12) %>% pull(tf_tissue)
```

```{r, fig.width = 12, fig.height = 12}

pp_top12 <- purrr::map(top12, function(tt){
    dt.ort_regr %>%
        dplyr::filter(tf_tissue == tt) %>% 
        ggplot(aes(x = abs(zscore), y = connectivity)) +
        geom_point() + theme_simple +
        labs(subtitle = tt, x = '|Z score|', y = 'GRN Connectivity')
})

pp_b <- patchwork::wrap_plots(pp_top12, nrow = 4, ncol = 3) + patchwork::plot_layout(axis_titles = "collect")


ggsave(file.path(figures_directory, 'GRN_concordance.scatterplot.absZscore_connectivity.png'), pp_b, width = 6500, height = 6500, dpi = 800, units = 'px')

pp_b

```


This is weird. It seems as if the edge weights are bimodal. Some are positive, others are negative. I need to see if TFXcan replicates this. 

```{r, fig.width = 12, fig.height = 16}
pp_boxtop12 <- purrr::map(top12, function(tp){
    dt.ort_regr %>%
        dplyr::filter(tf_tissue == tp) %>%
        dplyr::mutate(grp = ifelse(connectivity > 0, 'pos', 'neg')) %>%
        ggplot(aes(x = grp, y = abs(zscore), fill = grp)) +
        geom_boxplot(col = 'black', outlier.shape = NA) +
        geom_jitter(alpha = 0.1, color = 'grey', size = 0.2) +
        labs(x = 'GRN Connectivity', y = '|TFXcan Z-score|', subtitle = tp) + theme_simple
})

pp_b <- patchwork::wrap_plots(pp_boxtop12, nrow = 4, ncol = 3) + patchwork::plot_layout(axis_titles = "collect")


ggsave(file.path(figures_directory, 'GRN_concordance.boxplot.absZscore_connectivity.png'), pp_b, width = 9000, height = 9000, dpi = 800, units = 'px')

pp_b
```


### Enrichment in relevant tissues vs. others

I carried out this test in Blood tissues. I can test if the p values are enriched in Blood vs other tissues.

```{r}
pval.blood <- result.grn_concordance %>% dplyr::filter(tissue %in% c('Blood', 'BoneMarrow', 'Thymus', "Cordblood", "LymphNode")) %>% dplyr::pull(pvalue)
pval.others <- result.grn_concordance %>% dplyr::filter(!tissue %in% c('Blood', 'BoneMarrow', 'Thymus', "Cordblood", "LymphNode")) %>% dplyr::pull(pvalue)
```

```{r}
ppqq <- ggplot_qqplot(pval.others, pval.blood, pch = NULL) +
    geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed') +
    labs(x = expression(-log[10] * 'P (Other Tissues)'), y = expression(-log[10] * 'P (Blood, Bone Marrow, Thymus, Lymph)'), subtitle = "") +
    theme_simple

ggsave(file.path(figures_directory, 'GRN_concordance.enrichment.blood_vs_other_tissues.QQplot.png'), ppqq, width = 3000, height = 3000, dpi = 800, units = 'px')

ppqq
```

Sadly, there is no enrichment in blood vs. other tissues.

# Does variation in TF gene expression correlate with binding-expression?

## Data preparation and correlation

```{r}
# Collect TFs from TFXcan
tfxcan_tfs <- '/beagle3/haky/users/temi/projects/TFPred-snakemake/data/ENPACT_CHIPPEAKS_2026-01-02/statistics/ENPACT_CHIPPEAKS_2026-01-02.compiled_stats.txt'
dt.tfxcan_tfs <- data.table::fread(tfxcan_tfs) %>% dplyr::filter(type == 'test')
head(dt.tfxcan_tfs); dim(dt.tfxcan_tfs)
```

```{r}
tss_metadata <- data.table::fread("/beagle3/haky/users/temi/projects/dgtex/files/canonical_tss.matched_genes.tsv")
# remove duplicates
dups <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::group_by(locus) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt)) %>% 
    dplyr::filter(cnt > 1)

dt.tss_metadata <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::filter(!locus %in% dups$locus) %>% 
    dplyr::distinct()
```

```{r}
# available.tfs <- available.tf_gene %>% dplyr::select(tf) %>% distinct()
# available.genes <- available.tf_gene %>% dplyr::select(gene) %>% dplyr::inner_join(dt.tss_metadata, by = c('gene' = 'locus')) %>% distinct()
```
I need to subset the individuals that I have TF binding data for

```{r, eval = F}
# tfxcan.geu.individuals <- "/scratch/midway3/temi/enpact_scores/geuvadis_tss_chippeaks_models"
# inds <- list.files(tfxcan.geu.individuals) %>% strsplit('\\.') %>% sapply(getElement, 1)

gexpr_processed <- data.table::fread('/beagle3/haky/data/Geuvadis/expression/GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz') %>%
    tidyr::separate_wider_delim(Gene_Symbol, delim = '.', names = c('gene_name', 'id')) %>% dplyr::select(-c(TargetID, id, Chr, Coord)) %>%
    dplyr::inner_join(dt.tss_metadata, by = c('gene_name' = 'ensembl_gene_id')) %>%
    dplyr::relocate(gene_name, locus, external_gene_name)

gexpr_processed <- gexpr_processed %>% dplyr::select(-locus)
gexpr_processed[1:5, ]; dim(gexpr_processed)
```

```{r, eval = F}
protcod_genes.expression <- gexpr_processed %>%
    dplyr::select(-external_gene_name)
dim(protcod_genes.expression)

# nontf_genes.expression <- gexpr_processed %>%
#     dplyr::filter(!external_gene_name %in% dt.tfxcan_tfs$assay) %>%
#     dplyr::select(-external_gene_name)
# dim(nontf_genes.expression)


length(protcod_genes.expression$gene_name) == length(unique(protcod_genes.expression$gene_name))

# length(nontf_genes.expression$gene_name) == length(unique(nontf_genes.expression$gene_name))

protcod_genes.expression %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.protein_coding_genes.matrixA.tsv.gz"), row.names = F, col.names = T, quote = F, sep = '\t', compress = 'gzip')

protcod_genes.expression %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.protein_coding_genes.matrixB.tsv.gz"), row.names = F, col.names = T, quote = F, sep = '\t', compress = 'gzip')
```


```{r}
# test.expr <- tf_genes.expression %>% dplyr::filter(gene_name %in% c('ENSG00000152284', 'ENSG00000094631', 'ENSG00000130382'))

# test.expr %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.test.matrix.tsv.gz"), row.names = T, col.names = T, quote = F, sep = '\t', compress = 'gzip')
```

This correlates tf expression with target gene expression
```{r}
matA <- file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.protein_coding_genes.matrixA.tsv.gz")
matB <- file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.protein_coding_genes.matrixB.tsv.gz")
matTEMP <- "/scratch/beagle3/temi/gene_modules/coexpression/temp" #"/scratch/beagle3/temi/gene_modules/correlations/temp"
dir.create(matTEMP, recursive = TRUE)
matRES <- file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.co_expression.protein_coding_genes.correlations.tsv.gz")
```
```{r}
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/correlate_binding_expression.acrs.sbatch {matA} {matB} {matTEMP} {matRES}")
```


## Analysis of results
Here I will merge with the TFXcan promoter results and remove nearby, cis-regulators.

```{r}
dt.corr.expr.tf_nontf <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.tf_vs_nontf_genes.correlations.tsv.gz"))
dim(dt.corr.expr.tf_nontf); dt.corr.expr.tf_nontf[1:5, ]
```

```{r, cache.lazy = FALSE}

if(!base::exists("corr.expr_binding")){
    corr.expr_binding <- data.table::fread('/scratch/midway3/temi/enpact_scores/expression_vs_binding.enpact_scores.geuvadis.chippeak_models.correlation.tsv.gz') %>%
    dplyr::rename(tf_tissue = tf) %>%
    tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F)
    dim(corr.expr_binding); corr.expr_binding[1:5, ]
} else {
    dim(corr.expr_binding); corr.expr_binding[1:5, ]
}

```

```{r}
summary(corr.expr_binding$spearman_corr); summary(abs(corr.expr_binding$spearman_corr)); summary(dt.corr.expr.tf_nontf$corr_coef); summary(abs(dt.corr.expr.tf_nontf$corr_coef))
```

```{r, eval = F}
tss_metadata <- data.table::fread("/beagle3/haky/users/temi/projects/dgtex/files/canonical_tss.matched_genes.tsv")
# remove duplicates
dups <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::group_by(locus) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt)) %>% 
    dplyr::filter(cnt > 1)

dt.tss_metadata <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::filter(!locus %in% dups$locus) %>% 
    dplyr::distinct()

#
gene_info <- dt.tss_metadata %>% dplyr::select(ensembl_gene_id, locus)

#
gene_info.2 <- dt.tss_metadata %>% dplyr::select(ensembl_gene_id, external_gene_name)
```

For this analysis I will focus on blood TFs in TFXcan

```{r, eval = F}
# , 'BoneMarrow', "Cordblood"
blood.expr_binding <- corr.expr_binding %>%
    dplyr::filter(tissue %in% c('Blood')) %>%
    dplyr::rename(pvalue = spearman_pvalue)

blood.expr_binding[1:5, ]; dim(blood.expr_binding)

# 

mod.corr.expr_binding <- dplyr::inner_join(blood.expr_binding, gene_info, by = c('gene' = 'locus')) %>%
    dplyr::left_join(gene_info.2, by = c('tf' = 'external_gene_name'), suffix = c('.target_gene', '.tf_gene')) %>%
    dplyr::select(ensembl_gene_id.target_gene, ensembl_gene_id.tf_gene, pvalue, spearman_corr) %>%
    dplyr::rename(target_gene = ensembl_gene_id.target_gene, tf_gene = ensembl_gene_id.tf_gene)
mod.corr.expr_binding[1:5, ]

# Merge or join with the TF gene-Target gene data
dt.binding_expression.concordance <- dplyr::inner_join(dt.corr.expr.tf_nontf, mod.corr.expr_binding, by = c('target_gene', 'tf_gene'), suffix = c('.expr', '.binding')) %>% dplyr::rename(corr.expr = corr_coef, corr.binding = spearman_corr)


dt.binding_expression.concordance[1:5, ]; dim(dt.binding_expression.concordance)

```

### Remove nearby genes

```{r, eval = F}
trans_targets <- data.table::fread('/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/downloads/DatasetS1_trans-PrediXcan_whole_blood_FHS_results.txt.gz') %>% dplyr::select(trans_gene = predgene, target_gene = obsgene, FHS_beta, FHS_pval)
trans_targets[1:3, ]; dim(trans_targets)
```

Select for these trans targets. To do this correctly, I should select target genes for each trans gene and use that as a basis to select within the co-expression/binding results. 
```{r}
# dt_results <- dt.binding_expression.concordance %>%
#     dplyr::filter(tf_gene %in% trans_targets$trans_gene & target_gene %in% trans_targets$target_gene)

# dt_results[1:5, ]; dim(dt_results)
```

```{r, eval = F}
tfM <- dt.binding_expression.concordance %>%
    dplyr::filter(tf_gene %in% trans_targets$trans_gene)

# split the coexpression binding data by tf gene
tfM_split <- base::split(tfM, f = tfM$tf_gene)

# split the predixcan data by trans gene
pxM_split <- trans_targets %>% dplyr::filter(trans_gene %in% names(tfM_split)) %>%
    base::split(., f = .$trans_gene)

# ensure that the names are equally ordered
cnames <- intersect(names(tfM_split), names(pxM_split))
tfM_split <- tfM_split[cnames]
pxM_split <- pxM_split[cnames]
```

```{r, eval = F}
length(names(tfM_split)); length(names(pxM_split))
```

```{r, eval = F}
sapply(pxM_split, nrow)
```

```{r, eval = F}
# for each tf_gene, which is also a trans_gene/target, select the target genes within the tfM data
select.tfM_split <- purrr::map(cnames, function(cn){
    # tfM_split[[cn]] %>%
    #     dplyr::filter(target_gene %in% pxM_split[[cn]]$target_gene)
    tfM_split[[cn]] %>%
        dplyr::inner_join(pxM_split[[cn]], by = c('tf_gene' = 'trans_gene', 'target_gene'))
}) %>%
    dplyr::bind_rows()

select.tfM_split %>%
    data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "coexpression_binding_transregulation.data.tsv.gz"), row.names = F, col.names = T, quote = F, sep = '\t', compress = 'gzip')
```

### First, analyze the coexpression results
#### What is 5-point summary stats of the p values and correlation coefficients?

```{r}
summary(dt.corr.expr.tf_nontf$corr_coef); summary(dt.corr.expr.tf_nontf$pvalue)
```


### What is the distribution of the correlation coefficients?

```{r}
dt.corr.expr.tf_nontf$corr_coef |> summary()
```

```{r}
pp.all_corr <- ggplot() +
    geom_histogram(aes(x = dt.corr.expr.tf_nontf$corr_coef), fill = 'grey', col = 'black') +
    theme_simple +
    labs(x = 'Correlation Coefficient', y = 'TF gene-Target gene pair', subtitle = 'Correlation of\ngene expression')
pp.all_corr
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.coefficients.histogram.png"), pp.all_corr, width = 2000, height = 2000, units = 'px', dpi = 800)
```

```{r}
pp.all_corr.2 <- ggplot() +
    geom_boxplot(aes(y = dt.corr.expr.tf_nontf$corr_coef), fill = 'grey', col = 'black') +
    theme_simple +
    labs(y = 'Correlation Coefficient', x = '', subtitle = 'Correlation of\ngene expression') +
    theme(
        axis.title.x = element_blank(), # Removes the x-axis title
        axis.text.x = element_blank(),  # Removes the tick labels
        axis.ticks.x = element_blank()  # Removes the tick marks
    )

pp.all_corr.2
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.coefficients.boxplot.png"), pp.all_corr.2, width = 2000, height = 2000, units = 'px', dpi = 800)
```

::: {.callout}
There is high correlation between expression of some TFs and non-TF genes
:::

### Median correlations

```{r}
med_correlations <- dt.corr.expr.tf_nontf %>%
    dplyr::group_by(tf_gene) %>%
    dplyr::summarize(medCorr = median(corr_coef, na.rm = TRUE))
```

```{r}
pp.median_corr <- ggplot() +
    geom_boxplot(aes(y = med_correlations$medCorr), fill = 'grey', col = 'black') +
    theme_simple +
    labs(y = 'Correlation Coefficient', x = '', subtitle = 'Correlation of\ngene expression') +
    theme(
        axis.title.x = element_blank(), # Removes the x-axis title
        axis.text.x = element_blank(),  # Removes the tick labels
        axis.ticks.x = element_blank()  # Removes the tick marks
    )

pp.median_corr
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.median_coefficients.boxplot.png"), pp.median_corr, width = 2000, height = 2000, units = 'px', dpi = 800)
```

### Are the p values significant?

```{r}
pp.all_pvalues <- qqunif.ggplot(dt.corr.expr.tf_nontf$pvalue, pch = '.') + theme_simple
pp.all_pvalues
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.QQplot.pvalues.png"), pp.all_pvalues, width = 2000, height = 2000, units = 'px', dpi = 800)
```

### Using ACAT to aggregate the p values

```{r}
perTFGeneAcat <- dt.corr.expr.tf_nontf %>%
    dplyr::group_by(tf_gene) %>%
    dplyr::summarize(acatP = ACAT::ACAT(pvalue))
perTFGeneAcat[1:5, ]
```

```{r}
pp.acat_pvalues <- qqunif.ggplot(perTFGeneAcat$acatP, pch = '.') + theme_simple
pp.acat_pvalues
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.QQplot.acat_pvalues.png"), pp.acat_pvalues, width = 2000, height = 2000, units = 'px', dpi = 800)
```

::: {.callout}
These correlations are significant
:::

### Is every TF-gene correlated with at least one non-TF-gene?

```{r}
perGeneMaxCorr <- dt.corr.expr.tf_nontf %>%
    dplyr::group_by(tf_gene) %>%
    dplyr::summarize(maxCorr = max(abs(corr_coef), na.rm = TRUE))
perGeneMaxCorr[1:5, ]
```

```{r}
pp.max_corr <- ggplot() +
    geom_histogram(aes(x = perGeneMaxCorr$maxCorr), fill = 'grey', col = 'black') +
    theme_simple +
    labs(x = 'Max Correlation Coefficient\nper TF gene', y = 'Number of TF genes', subtitle = 'Max Correlation of gene\nexpression per TF gene')
pp.max_corr 
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.max_per_tf_corr.histogram.png"), pp.max_corr, width = 2000, height = 2000, units = 'px', dpi = 800)
```


### Using pi1 values, what is the proportion of truly correlated gene per TF-gene?
```{r}
pi1_estimator <- function(x){
    x <- x[x != 0]
    1 - qvalue::qvalue(x)$pi0
}

perTFGenePi1 <- dt.corr.expr.tf_nontf %>%
    dplyr::group_by(tf_gene) %>%
    dplyr::summarize(pi1 = pi1_estimator(pvalue))
perTFGenePi1[1:5, ]
```

```{r}
pp.pi1 <- ggplot() +
    geom_histogram(aes(x = perTFGenePi1$pi1), fill = 'grey', col = 'black') +
    theme_simple +
    labs(x = 'pi1', y = 'Number of TF genes', subtitle = 'Proportion of truly\ncorrelated genes per TF gene')
pp.pi1
```

```{r, eval = F}
ggsave(file.path(figures_directory, "coexpression.correlation.pi1.histogram.png"), pp.pi1, width = 2000, height = 2000, units = 'px', dpi = 800)
```

### Can I show single examples?

```{r}
tf_genes.expression <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.tf_genes.matrix.tsv.gz")) %>% tibble::column_to_rownames('gene_name') %>% as.matrix()

nontf_genes.expression <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.nontf_genes.matrix.tsv.gz")) %>% tibble::column_to_rownames('gene_name') %>% as.matrix()

coln <- intersect(colnames(tf_genes.expression), colnames(nontf_genes.expression))

tf_genes.expression <- tf_genes.expression[, coln]
nontf_genes.expression <- nontf_genes.expression[, coln]

tf_genes.expression[1:5, 1:5]; nontf_genes.expression[1:5, 1:5]
```

```{r}
# top 12 most correlated
top12 <- dt.corr.expr.tf_nontf %>%
    dplyr::arrange(desc(abs(corr_coef))) %>%
    head(n = 12)
```

```{r, fig.width = 12, fig.height = 12}
p12 <- purrr::map(1:nrow(top12), function(i){
    temp <- top12[i, ]
    ggplot() +
        geom_point(aes(x = tf_genes.expression[temp$tf_gene, ], y = nontf_genes.expression[temp$target_gene, ])) +
        theme_simple +
        labs(x = glue("{temp$tf_gene}, TF"), y = glue("{temp$target_gene}, Target Gene"))
}, .progress = TRUE)

ppw.p12 <- patchwork::wrap_plots(p12, nrow = 4, ncol = 3) + patchwork::plot_layout(axis_titles = "collect")
ppw.p12
```


```{r, eval = F}
ggsave(file.path(figures_directory, 'coexpression.scatterplot.examples.png'), ppw.p12, width = 6500, height = 6500, dpi = 800, units = 'px')
```

## Analysis of the coexpression, binding, and transregulation results after removing nearby genes

```{r}
dt_results <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "coexpression_binding_transregulation.data.tsv.gz"))
dim(dt_results); dt_results[1:3, ]
```

```{r}
# merge with metadata to add gene names)
mtd <- dt.tss_metadata %>% dplyr::select(ensembl_gene_id, external_gene_name)

dt_results <- dt_results %>%
    dplyr::inner_join(mtd, by = c('tf_gene' = 'ensembl_gene_id')) %>%
    dplyr::rename(tf_gene_name = external_gene_name) %>%
    dplyr::inner_join(mtd, by = c('target_gene' = 'ensembl_gene_id')) %>%
    dplyr::rename(target_gene_name = external_gene_name)

dt_results[1:3, ]
```

### Correlate with promoter-TFXcan

Does predicted variability in TF gene-Target gene expression correlate with TF binding-Target gene?

```{r}
# how many of these do I have left?
dt_results$tf_gene |> unique() |> length()
dt_results$corr.expr |> summary()
dt_results$corr.binding |> summary()
```

How are the p values distributed?

```{r, fig.width = 12}
qq_expr <- qqunif.ggplot(p = dt_results$pvalue.expr) + theme_simple + labs(subtitle = 'Coexpression')

qq_binding <- qqunif.ggplot(p = dt_results$pvalue.binding) + theme_simple + labs(subtitle = 'Binding')

qq_transreg <- qqunif.ggplot(p = dt_results$FHS_pval) + theme_simple + labs(subtitle = 'Transregulation')

ppw.qqthree <- patchwork::wrap_plots(qq_expr, qq_binding, qq_transreg, nrow = 1, ncol = 3) + patchwork::plot_layout(axis = "collect_y")
ppw.qqthree
```

```{r, eval = F}
ggsave(file.path(figures_directory, 'coexpression_binding_transregulation.QQplots.png'), ppw.qqthree, width = 6500, height = 2500, dpi = 800, units = 'px')
```

### How are these split per TF gene?

If a TF is significantly coexpressed with a gene, is coregulation significant too?

```{r}
list.concordance <- base::split(dt_results, f = dt_results$tf_gene_name)
length(list.concordance)
```

```{r}
list.concordance[[1]]$pvalue.binding |> hist()
```

```{r}
list.concordance[[1]]$pvalue.expr |> hist()
```


```{r}
pvs <- purrr::map(names(list.concordance), function(lc){
    with(list.concordance[[lc]], cor.test(x = abs(pvalue.expr), y = abs(pvalue.binding)))$p.value
}) %>% unlist()
```

```{r}
hist(pvs)
```

Correlation of co-expression and binding for each TF-gene, absolute values
```{r, eval = F}
plots_abs_corr <- purrr::map(names(list.concordance), function(lc){
    ggplot(list.concordance[[lc]], aes(x = abs(corr.expr), y = abs(corr.binding))) +
    geom_point() +
    theme_simple + 
    geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed') +
    labs(x = '|correlation of coexpression|', y = '|correlation of binding|', subtitle = glue::glue("{lc}: coexpression vs binding"))
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'coexpression_vs_binding.abs_correlations.pdf'), gridExtra::marrangeGrob(grobs = plots_abs_corr, nrow=1, ncol=1))
```


Correlation of co-expression and binding for each TF-gene, raw values
```{r, eval = F}
plots_corr <- purrr::map(names(list.concordance), function(lc){
    ggplot(list.concordance[[lc]], aes(x = corr.expr, y = corr.binding)) +
    geom_point() +
    theme_simple +
    geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed') +
    labs(x = 'correlation of expression', y = 'correlation of binding', subtitle = glue::glue("{lc}: coexpression vs binding"))
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'coexpression_vs_binding.correlations.pdf'), gridExtra::marrangeGrob(grobs = plots_corr, nrow=1, ncol=1))
```
qqplot, comparing each p values against the null
```{r, eval = F}
# purrr and plot
plotlist <- purrr::map(names(list.concordance), function(lc){
    tgene <- list.concordance[[lc]] %>%
        dplyr::filter(pvalue.expr != 0, pvalue.binding != 0)

    qqcompare.ggplot(p1 = tgene$pvalue.expr, p2 = tgene$pvalue.binding, names = c('Coexpression', 'Binding')) + theme_simple + scale_color_manual(name = '', values = c('Coexpression' = 'red', 'Binding' = 'darkgreen')) +
    labs(subtitle = glue('{lc}: QQ plot of p values'))

    # ggplot_qqplot(x = tgene$pvalue.tfxcan, y = tgene$pvalue.gtex) +
    #     geom_abline(intercept = 0, slope = 1, col = 'red') +
    #     theme_simple +
    #     labs(x = 'TFXcan', y = 'GTEx', subtitle = glue('{cll}: QQ plot of p values'))
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'coexpression_vs_binding.QQplot.pvalues.pdf'), gridExtra::marrangeGrob(grobs = plotlist, nrow=1, ncol=1))
```

```{r, eval = F}
plotlist <- purrr::map(names(list.concordance), function(lc){
    tgene <- list.concordance[[lc]] %>%
        dplyr::filter(pvalue.expr != 0, pvalue.binding != 0)

    tgene %>%
        ggplot(aes(-log10(pvalue.expr), -log10(pvalue.binding))) + 
        geom_point() +
        theme_simple +
        labs(subtitle = glue('{lc}: -log10(p values)')) +
        geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed')
}, .progress = TRUE)


ggsave(file.path(figures_directory, 'coexpression_vs_binding.neglog10.pvalues.pdf'), gridExtra::marrangeGrob(grobs = plotlist, nrow=1, ncol=1))
```

### Compare coexpression with binding

```{r, eval = F}
plotlist <- purrr::map(names(list.concordance), function(lc){
    with(list.concordance[[lc]], ggplot_qqplot(-log10(pvalue.expr), -log10(pvalue.binding))) +
        theme_simple +
        labs(x = "Coexpression", y = 'Binding', subtitle = glue('{lc}: QQ plot -log10(p values)')) +
        geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed')
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'coexpression_vs_binding.QQplot.neglog10.pvalues.pdf'), gridExtra::marrangeGrob(grobs = plotlist, nrow=1, ncol=1))
```

### Compare coexpression with transregulation

```{r, eval = F}
plotlist <- purrr::map(names(list.concordance), function(lc){
    with(list.concordance[[lc]], ggplot_qqplot(-log10(FHS_pval), -log10(pvalue.expr))) +
        theme_simple +
        labs(x = "Transregulation", y = 'Coexpression', subtitle = glue('{lc}: QQ plot -log10(p values)')) +
        geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed')
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'coexpression_vs_transregulation.QQplot.neglog10.pvalues.pdf'), gridExtra::marrangeGrob(grobs = plotlist, nrow=1, ncol=1))
```

### Compare binding with transregulation
```{r, eval = F}
plotlist <- purrr::map(names(list.concordance), function(lc){
    with(list.concordance[[lc]], ggplot_qqplot(-log10(FHS_pval), -log10(pvalue.binding))) +
        theme_simple +
        labs(x = "Transregulation", y = 'Binding', subtitle = glue('{lc}: QQ plot -log10(p values)')) +
        geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed')
}, .progress = TRUE)

ggsave(file.path(figures_directory, 'binding_vs_transregulation.QQplot.neglog10.pvalues.pdf'), gridExtra::marrangeGrob(grobs = plotlist, nrow=1, ncol=1))
```


### Bonferroni adjustment
```{r}
num_bonferroni <- purrr::map(list.concordance, function(dt){
    dt %>%
        dplyr::mutate(ln = nrow(.), nexpr_signif = pvalue.expr < (0.05/ln), binding_signif = pvalue.binding < (0.05/ln)) %>%
        dplyr::group_by(tf_gene) %>%
        dplyr::summarize(num_expr_signif = sum(nexpr_signif == TRUE), num_binding_signif = sum(binding_signif == TRUE))
}, .progress = TRUE) %>%
    dplyr::bind_rows()
```

```{r}
num_bonferroni$num_expr_signif |> summary();
num_bonferroni$num_binding_signif |> summary()
```

- Aggregate using ACAT

```{r}
pi1list <- purrr::map(names(list.concordance), function(lc){
    list.concordance[[lc]] %>%
        dplyr::group_by(tf_gene_name) %>%
        dplyr::summarize(pi1.expr = 1 - qvalue::qvalue(pvalue.expr)$pi0,
            pi1.binding = 1 - qvalue::qvalue(pvalue.binding)$pi0)
}, .progress = TRUE) %>%
    dplyr::bind_rows()
```

```{r}
ppn_pi1 <- pi1list %>%
    ggplot(aes(x = pi1.expr, y = pi1.binding)) +
    geom_point() +
    theme_simple +
    labs(x = "Expression (pi1)", y = 'Binding (pi1)', subtitle = 'Proportion of truly correlated\n(binding or expression) \nTF-target genes') +
    geom_abline(intercept = 0, slope = 1, col = 'grey', linetype = 'dashed')
ppn_pi1
```

```{r, eval = F}
ggsave(file.path(figures_directory, 'coexpression_vs_binding.pi1_values.png'), ppn_pi1, width = 2000, height = 2000, dpi = 800, units = 'px')
```

### Can co-expression or binding predict transregulation?

```{r}
library(PRROC)
label <- as.integer(dt_results$FHS_pval < 0.05)
roc_expr <- PRROC::roc.curve(scores.class0 = -log10(dt_results$pvalue.expr), weights.class0 = label, curve = TRUE)
roc_binding <- PRROC::roc.curve(scores.class0 = -log10(dt_results$pvalue.binding), weights.class0 = label, curve = TRUE)

plot(roc_expr);    roc_expr$auc
plot(roc_binding); roc_binding$auc

```

### Are the most significant transregulators significant in coexpression or binding?

```{r}
# qqcompare.ggplot(
#   p1 = dt_results$pvalue.expr,
#   p2 = dt_results$FHS_pval,
#   names = c('Expression', 'Transregulation')
# ) + theme_simple

```

```{r}
pi1_expr <- 1 - qvalue::qvalue(dt_results$pvalue.expr)$pi0
pi1_transreg <- 1 - qvalue::qvalue(dt_results$FHS_pval)$pi0
pi1_binding <- 1 - qvalue::qvalue(dt_results$pvalue.binding)$pi0
cat("pi1 expression:      ", pi1_expr, "\n")
cat("pi1 transregulation: ", pi1_transreg, "\n")
cat("pi1 binding: ", pi1_binding, "\n")
```

```{r}
pwtest <- wilcox.test(
  -log10(dt_results$pvalue.expr),
  -log10(dt_results$FHS_pval),
  paired = TRUE,
  alternative = "greater"  # test if expression is more enriched
)

pwtest$p.value
```

```{r}
cutoff <- 0.05
expr_sig <- dt_results$pvalue.expr < (cutoff/nrow(dt_results))
transreg_sig <- dt_results$FHS_pval < (cutoff/nrow(dt_results))

tab <- table(expr_sig, transreg_sig)
tab
mcnemar.test(tab)  # tests asymmetry: are pairs where only expr is sig > pairs where only transreg is sig?
```

```{r}
cutoff <- 0.05
binding_sig <- dt_results$pvalue.binding < (cutoff/nrow(dt_results))
transreg_sig <- dt_results$FHS_pval < (cutoff/nrow(dt_results))

tab <- table(binding_sig, transreg_sig)
tab
mcnemar.test(tab)  # tests asymmetry: are pairs where only expr is sig > pairs where only transreg is sig?
```


```{r}
# library(mediation)
# # For one TF-target pair:
# tf_expr    <- tf_genes.expression["ENSG...", ]
# binding    <- # enpact score vector for this TF-target pair
# target_expr <- nontf_genes.expression["ENSG...", ]

# df <- data.frame(tf_expr, binding, target_expr)
# m.mediator <- lm(binding ~ tf_expr, data = df)
# m.outcome  <- lm(target_expr ~ tf_expr + binding, data = df)
# med <- mediation::mediate(m.mediator, m.outcome, treat = "tf_expr", mediator = "binding")
# summary(med)

```

### Jaccard clustering of TF target gene sets

```{r}
sig_targets <- dt_results %>%
    dplyr::filter(FHS_pval < 0.05) %>%
    base::split(., .$tf_gene_name) %>%
    purrr::map(~ .$target_gene)

# Pairwise Jaccard
jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

tfs <- names(sig_targets)
jmat <- outer(tfs, tfs, Vectorize(function(i, j) jaccard(sig_targets[[i]], sig_targets[[j]])))
dimnames(jmat) <- list(tfs, tfs)

ComplexHeatmap::Heatmap(jmat, col = brewer.pal(n = 8, name = "Reds"), cluster_rows = T, cluster_columns = T, column_names_rot = 90, 
        heatmap_legend_param = list(title = expression("Similarity")), row_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'), column_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'))

```

```{r}
sig_targets <- dt_results %>%
    dplyr::filter(pvalue.expr < (cutoff/nrow(dt_results))) %>%
    base::split(., .$tf_gene_name) %>%
    purrr::map(~ .$target_gene)

# Pairwise Jaccard
jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

tfs <- names(sig_targets)
jmat <- outer(tfs, tfs, Vectorize(function(i, j) jaccard(sig_targets[[i]], sig_targets[[j]])))
dimnames(jmat) <- list(tfs, tfs)

ComplexHeatmap::Heatmap(jmat, col = brewer.pal(n = 8, name = "Reds"), cluster_rows = T, cluster_columns = T, column_names_rot = 90, 
        heatmap_legend_param = list(title = expression("Similarity")), row_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'), column_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'))
```

```{r}
sig_targets <- dt_results %>%
    dplyr::filter(pvalue.binding < 0.05) %>%
    base::split(., .$tf_gene_name) %>%
    purrr::map(~ .$target_gene)

# Pairwise Jaccard
jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))

tfs <- names(sig_targets)
jmat <- outer(tfs, tfs, Vectorize(function(i, j) jaccard(sig_targets[[i]], sig_targets[[j]])))
dimnames(jmat) <- list(tfs, tfs)

ComplexHeatmap::Heatmap(jmat, col = brewer.pal(n = 8, name = "Reds"), cluster_rows = T, cluster_columns = T, column_names_rot = 90, 
        heatmap_legend_param = list(title = expression("Similarity")), row_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'), column_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'))
```


### Do TFs in the same family tend to bind to the same overlapping genes?

```{r}
library(openxlsx)
```

```{r}
tf_family <- openxlsx::read.xlsx('/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/downloads/1-s2.0-S0092867418301065-mmc2.xlsx', sheet = 2, startRow = 2) %>% dplyr::filter(X4 == 'Yes') %>% dplyr::select(ID, Name, DBD) %>%
    dplyr::filter(DBD != 'Unknown')
```

```{r}
dim(tf_family); tf_family[1:5, ]
```

```{r}
# Use FHS transregulation as "regulated targets" (most stringent)
sig_targets <- dt_results %>%
    dplyr::filter(pvalue.binding < (0.05/nrow(dt_results))) %>%
    base::split(., .$tf_gene_name) %>%
    purrr::map(~ .$target_gene)

tfs <- names(sig_targets)
jaccard <- function(a, b) length(intersect(a, b)) / length(union(a, b))
jmat <- outer(tfs, tfs, Vectorize(function(i, j) jaccard(sig_targets[[i]], sig_targets[[j]])))
dimnames(jmat) <- list(tfs, tfs)
```

```{r}
# Long-format pairwise Jaccard (upper triangle only)
jlong <- as.data.frame(as.table(jmat)) %>%
    setNames(c('tf1', 'tf2', 'jaccard')) %>%
    dplyr::filter(as.character(tf1) != as.character(tf2))  # avoid duplicates

# Join family annotation for both TFs
jlong <- jlong %>%
    dplyr::inner_join(tf_family, by = c('tf1' = 'Name')) %>%
    dplyr::rename(family1 = DBD) %>%
    dplyr::inner_join(tf_family, by = c('tf2' = 'Name')) %>%
    dplyr::rename(family2 = DBD) %>%
    dplyr::mutate(same_family = family1 == family2) %>%
    dplyr::filter(!is.na(same_family)) %>% 
    dplyr::filter(tf1 != tf2)
```

```{r}
# jlong %>% dplyr::filter(!is.na(same_family)) %>% dplyr::filter(tf1 != tf2) %>% dplyr::filter(same_family == TRUE)
```

```{r}
obs_diff <- with(jlong, mean(jaccard[same_family]) - mean(jaccard[!same_family]))

set.seed(2025)
perm_diffs <- replicate(1000, {
    shuffled <- jlong %>% dplyr::mutate(same_family = sample(same_family))
    with(shuffled, mean(jaccard[same_family]) - mean(jaccard[!same_family]))
})

perm_pval <- mean(perm_diffs >= obs_diff)
```

```{r}
hist(perm_diffs, xlab = 'Difference in Jaccard similarity', ylab = 'Number of TF regulators')
abline(v = obs_diff, col = "red", lwd = 2, lty = 2) 
```

```{r}
pp.jaccard.perm <- ggplot(jlong %>% dplyr::filter(!is.na(same_family)), 
       aes(x = same_family, y = jaccard, fill = same_family)) +
    geom_boxplot(outlier.shape = NA, col = 'black') +
    geom_jitter(alpha = 0.2, size = 0.6, color = 'grey') +
    theme_simple +
    scale_x_discrete(labels = c('Different family', 'Same family')) +
    labs(x = '', y = 'Jaccard similarity of target sets',
         subtitle = 'Do same-family TFs share more targets?') +
    theme(legend.position = 'none')


pp.jaccard.perm

ggsave(file.path(figures_directory, 'jaccard_TF_family.png'), pp.jaccard.perm, width = 3000, height = 3000, dpi = 800, units = 'px')
```

```{r}
jlong %>% dplyr::filter(!is.na(same_family)) %>% dplyr::filter(same_family == TRUE)
```

```{r}
wtest <- wilcox.test(jaccard ~ same_family, data = jlong %>% dplyr::filter(!is.na(same_family))) 
```

```{r}
wtest$p.value
```

```{r}
jlong %>% dplyr::filter(!is.na(same_family)) %>% dplyr::filter(same_family == TRUE)
```


4. Mediation: does binding mediate TF expression → target expression?
You have all three variables. Test if TF binding at the target TSS is on the causal path between TF gene expression and target gene expression using mediation:

```{r, eval = F}
library(mediation)
# For one TF-target pair:
tf_expr    <- tf_genes.expression["ENSG...", ]
binding    <- # enpact score vector for this TF-target pair
target_expr <- nontf_genes.expression["ENSG...", ]

df <- data.frame(tf_expr, binding, target_expr)
m.mediator <- lm(binding ~ tf_expr, data = df)
m.outcome  <- lm(target_expr ~ tf_expr + binding, data = df)
med <- mediation::mediate(m.mediator, m.outcome, treat = "tf_expr", mediator = "binding")
summary(med)
```


6. Pathway enrichment on concordant TF target sets
For TFs where all three signals agree (co-expression + binding + transregulation significant), test if their target genes are enriched for known pathways:

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r}
concordant_targets <- dt_results %>%
    dplyr::filter(pvalue.expr < 0.05, pvalue.binding < 0.05, FHS_pval < 0.05) %>%
    dplyr::pull(target_gene) %>% unique()

enrichment_result <- enrichGO(gene = concordant_targets, OrgDb = org.Hs.eg.db, keyType = "ENSEMBL", ont = "BP")
```

```{r}
enrichment_result
```

```{r}
dotplot(enrichment_result,showCategory=20,font.size=10,label_format=70)+
  scale_size_continuous(range=c(1, 7))+
  theme_minimal() +
  ggtitle("GO Enrichment of up-regulated genes")
```


### Split TFs into 4 kinds
1. Signif expression = signif binding
2. Signif expression != signif binding

```{r}
dt_bonferroni <- purrr::map(list.concordance, function(dt){
    dt %>%
        dplyr::mutate(ln = nrow(.), expr_signif = pvalue.expr < (0.05/ln), binding_signif = pvalue.binding < (0.05/ln))
}, .progress = TRUE) %>%
    dplyr::bind_rows()
```


### UpSet plot of TF-target pair significance across all three signals

```{r}
library(UpSetR)

# Build a per-TF-target membership matrix for all three signals
# Use Bonferroni threshold for expr and binding (as computed above),
# and nominal p < 0.05 for transregulation (FHS)
upset_input <- purrr::map(list.concordance, function(dt){
    dt %>%
        dplyr::mutate(
            ln             = nrow(.),
            expr_signif    = as.integer(pvalue.expr     < (0.05 / ln)),
            binding_signif = as.integer(pvalue.binding  < (0.05 / ln)),
            transreg_signif = as.integer(FHS_pval       < (0.05 / ln))
        )
}, .progress = TRUE) %>%
    dplyr::bind_rows() %>%
    dplyr::select(expr_signif, binding_signif, transreg_signif)

pp_upset <- UpSetR::upset(
    upset_input,
    sets            = c('expr_signif', 'binding_signif', 'transreg_signif'),
    sets.bar.color  = c('#E41A1C', '#377EB8', '#4DAF4A'),
    order.by        = 'freq',
    mainbar.y.label = 'Number of TF-target pairs',
    sets.x.label    = 'Total significant pairs',
    text.scale      = c(1.2, 1.1, 1, 1, 1.1, 0.9)
)

pp_upset
```

```{r, eval = F}
png(file.path(figures_directory, 'upset.expr_binding_transreg.significance.png'),
    width = 5500, height = 3500, res = 800)
pp_upset
dev.off()
```



```{r, cache = FALSE}
utils::sessionInfo()

knitr::knit_exit()
```


## Examples

```{r}
dt.binding_expression.concordance$corr_coef %>% summary();

dt.binding_expression.concordance$spearman_corr %>% summary()
```

```{r}
dt.binding_expression.concordance %>%
    dplyr::slice_max(corr_coef)
```

```{r}
dt.binding_expression.concordance %>%
    dplyr::slice_max(spearman_corr)
```

```{r}
perGeneCorrList <- purrr::map(names(list.concordance), function(lc){
    list.concordance[[lc]] %>%
        dplyr::group_by(tf_gene) %>%
        dplyr::summarize(corr = cor.test(corr_coef, spearman_corr)$estimate)
}, .progress = TRUE) %>%
    dplyr::bind_rows() %>%
    dplyr::arrange(desc(corr))
```

```{r}
ppn_corr <- perGeneCorrList %>%
    ggplot(aes(x = corr)) +
    geom_histogram(col = 'black') +
    theme_simple +
    labs(x = "Correlation Coefficient", y = 'Number of TFs', subtitle = 'Correlation')

ggsave(file.path(figures_directory, 'expression_vs_binding.correlation_of_correlations.png'), ppn_corr, width = 3000, height = 3000, dpi = 800, units = 'px')
```

::: {.callout}

The effect of binding on expression and effect of expression on expression are not well-correlated.
:::

```{r}
dt.binding_expression.concordance %>% dplyr::filter(corr_coef > 0.6 & spearman_corr > 0.6)
```

```{r}
perTFGeneAcat <- dt.binding_expression.concordance %>%
    dplyr::group_by(tf_gene) %>%
    dplyr::summarize(acatP = ACAT::ACAT(pvalue))
perTFGeneAcat[1:5, ]
```

```{r}
pp.acat_pvalues <- qqunif.ggplot(perTFGeneAcat$meanCorr) + theme_simple

ggsave(file.path(figures_directory, "tfGene_vs_targetGene.qqplot.acat_pvalues.png"), pp.acat_pvalues, width = 3000, height = 3000, units = 'px', dpi = 800)
```

```{r}
dt.binding_expression.concordance %>%
    dplyr::filter(pvalue.tfxcan != 0) %>% dplyr::slice_head(n = 1000) %>% with(., cor.test(-log10(pvalue.tfxcan), -log10(pvalue.gtex)))
```

```{r}
length(unique(dt.binding_expression.concordance$tf_gene))
```

```{r}
tgene <- dt.binding_expression.concordance %>%
    dplyr::filter(tf_gene == 'ENSG00000152284')
```

```{r}
pppvalues <- qqcompare.ggplot(p1 = tgene$pvalue.tfxcan, p2 = tgene$pvalue.gtex, names = c('TFXcan', 'GTEx')) + theme_simple + scale_color_manual(name = '', values = c('TFXcan' = 'darkgreen', 'GTEx' = 'red'))
pppvalues
# ggsave(file.path(figures_directory, '../qqplot.GATA1_ACRs.pvalues.png'), pppvalues, width = 3000, height = 3000, dpi = 800, units = 'px')

# pppvalues <- ggplot_qqplot(x = tgene$pvalue.tfxcan), y = tgene$pvalue.gtex)) +

#  +
# labs(x = 'TFXcan', y = 'GTEx', subtitle = 'QQ plot of p values')
```



```{r}
qqcompare.ggplot <- function(p1, p2, names = c('observed', 'null'), mlog10_p_thres=30)
```

```{r}
with(dt.binding_expression.concordance[1:1000, ], cor.test(-log10(pvalue.tfxcan), -log10(pvalue.gtex)))
```






```{r}
summary(dt.res.pvalues$pvalue)
```

```{r}
1 - qvalue::qvalue(dt.res.pvalues$pvalue)$pi0
```

```{r}
pppvalues <- ggplot_qqplot(y = regulated_genes$pvalue, x = non_regulated_genes$pvalue) +
geom_abline(intercept = 0, slope = 1, col = 'red') +
theme_simple +
labs(y = 'ACRs target genes', x = 'ACRs non-target genes', subtitle = 'QQ plot of p values')

ggsave(file.path(figures_directory, '../qqplot.GATA1_ACRs.pvalues.png'), pppvalues, width = 3000, height = 3000, dpi = 800, units = 'px')
```

```{r}
dt.binding_expression.concordance %>%
    dplyr::filter(pvalue.tfxcan == 0) %>% dplyr::slice_head(n = 1)
```

```{r}
unique(dt.binding_expression.concordance$tf_gene)[1:4]
```


```{r}
dt.binding_expression.concordance %>%
    dplyr::filter(tf_gene == 'ENSG00000152284', gene == 'ENSG00000140807')
```

```{r}
dt.res.merged %>%
    dplyr::filter(tf_gene == 'ENSG00000152284', gene == 'ENSG00000140807')
```

```{r}

comnam <- intersect(colnames(nontf_genes.expression), colnames(tf_genes.expression))

nx <- nontf_genes.expression %>% dplyr::filter(gene_name == 'ENSG00000140807') %>% dplyr::select(-all_of(c('gene_name'))) %>% as.matrix()

nx <- nx[, comnam[-1]]

ny <- tf_genes.expression %>% dplyr::filter(gene_name == 'ENSG00000152284') %>% dplyr::select(-all_of(c('gene_name', 'locus'))) %>% as.matrix()

ny <- ny[, comnam[-1]]
```

```{r}
length(nx); length(ny)
```

```{r}
cte <- cor.test(nx, ny)
cte$estimate; cte$p.value
```








```{r}
knitr::knit_exit()
```



# Enhancer-target gene

Enhancer's downloaded from: http://www.licpathway.net/ENdb/index.php

http://reggen.iiitd.edu.in:1207/ePerturbDB-html/downloads.php

```{r}
enhancers_directory <- '/beagle3/haky/data/enhancers'
```
```{r}
utils::download.file("http://www.licpathway.net/ENdb/download/enhancer_main.txt", destfile = file.path(enhancers_directory, 'enhancer_main.txt'))
```


```{r}
dt <- data.table::fread(file.path(enhancers_directory, 'enhancer_perturbation/enh_perturb.tsv'))

```


# Chromatin accessibility - Target genes association

```{r}
perturbation_directory <- '/beagle3/haky/users/temi/data/enhancers/perturbation'
```
What TFs and genes do I have?
TFs from Enpact
Genes from Geuvadis LCL expression
```{r}
corr.expr_binding <- data.table::fread('/scratch/midway3/temi/enpact_scores/expression.regression.enpact_scores.geuvadis.chippeak_models.tsv.gz') %>%
    dplyr::filter(term == 'binding') %>%
    dplyr::rename(estimate = Estimate, se = `Std. Error`, tvalue = `t value`, pvalue = `Pr(>|t|)`) %>%
    dplyr::mutate(zscore = estimate/se) %>%
    tidyr::separate_wider_delim(col = tf_tissue, names = c('tf', 'tissue'), delim = '_', cols_remove = F)
dim(corr.expr_binding); corr.expr_binding[1:5, ]
```

```{r}
tss_metadata <- data.table::fread("/beagle3/haky/users/temi/projects/dgtex/files/canonical_tss.matched_genes.tsv")
# remove duplicates
dups <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::group_by(locus) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt)) %>% 
    dplyr::filter(cnt > 1)

dt.tss_metadata <- tss_metadata %>%
    dplyr::select(ensembl_gene_id, locus, external_gene_name) %>%
    dplyr::filter(!locus %in% dups$locus) %>% 
    dplyr::distinct()
```

```{r}
available.tf_gene <- corr.expr_binding %>%
    dplyr::select(tf, gene) %>% distinct()

available.tfs <- available.tf_gene %>% dplyr::select(tf) %>% distinct()
available.genes <- available.tf_gene %>% dplyr::select(gene) %>% dplyr::inner_join(dt.tss_metadata, by = c('gene' = 'locus')) %>% distinct()
```

```{r}
available.genes <- data.table::fread('/beagle3/haky/data/Geuvadis/expression/GD462.GeneQuantRPKM.50FN.samplename.resk10.txt.gz') %>%
    dplyr::select(TargetID, Gene_Symbol, Chr, Coord) %>%
    tidyr::separate_wider_delim(Gene_Symbol, delim = '.', names = c('gene_name', 'id')) %>% dplyr::select(-c(TargetID, id, Chr, Coord)) %>%
    dplyr::inner_join(available.genes, by = c('gene_name' = 'ensembl_gene_id'))
available.genes[1:5, ]
```

```{r}
dt_genes <- data.table::fread(file.path(perturbation_directory, 'genes_tf.tsv'))
dt_accessibility <- data.table::fread(file.path(perturbation_directory, 'accessible_regions_tf.tsv')) 
```

```{r}
dt_perturbation <- dplyr::inner_join(dt_genes, dt_accessibility, by = 'perturbation_name', suffix = c('.gene', '.accessibility')) 
dim(dt_perturbation)
```


```{r}
dt_genes %>%
    dplyr::group_by(perturbation_name) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt))
```

```{r}
dt_accessibility %>%
    dplyr::group_by(perturbation_name) %>%
    dplyr::summarize(cnt = n()) %>%
    dplyr::arrange(desc(cnt))
```

```{r}
dt_perturbation.large <- dt_perturbation %>% dplyr::filter(abs(log2FC.gene) > 1.5, abs(log2FC.accessibility) > 1.5)
dim(dt_perturbation.large)
```

```{r}
dt_perturbation.large[1:5, ]
```

Pick one TF: GATA1

Select all accessibility (ACR) and target genes
```{r}
GATA1.pert <- dt_perturbation.large %>% dplyr::filter(perturbation_name == 'GATA1') %>% dplyr::filter(gene_ID %in% available.genes$external_gene_name)
GATA1.pert[1:5, ]; dim(GATA1.pert)
```

```{r}
GATA1.pert$log2FC.gene |> summary()
GATA1.pert$log2FC.accessibility |> summary()
```


```{r}
# prepare the data
GATA1.pert.acrs <- GATA1.pert %>% dplyr::select(peak_locus) %>% dplyr::mutate(peak_locus = gsub("-", '_', peak_locus)) %>% distinct() %>% tidyr::separate_wider_delim(peak_locus, delim = '_', names = c('chrom', 'start', 'end'), cols_remove = FALSE) %>% dplyr::mutate(across(c('start', 'end'), as.numeric)) %>% dplyr::mutate(width = end - start)


GATA1.pert.acrs %>% dplyr::select(peak_locus) %>%
    data.table::fwrite(file.path(tfxcan_directory, 
    'experiments/gene_modules/data/GATA1.ACRs.tsv'), col.names = F, row.names = F, quote = F, sep = '\t')
```

```{r}
GATA1.pert.acrs$width |> summary()
```


## Predict epigenomic features and TF binding
```{r}
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/enformer_predict.sbatch /beagle3/haky/users/temi/projects/TFXcan-snakemake/workflow/enformer/enformer_predict.py /beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/data/GATA1.ACRs.enformer_params.yaml")
```

```{r}
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/enformer_merge.sbatch /beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/data/GATA1_ACRs.aggregation_config.yaml /scratch/beagle3/temi/gene_modules/epigenomes GATA1_ACRs.epigenomes.h5")
```

```{r}
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/enformer_process.sbatch /scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.h5 /scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed")
```

```{r}
# /beagle3/haky/users/temi/projects/TFXcan/src/enpact_predict.sbatch
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/enpact_predict.sbatch /scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.matrix.h5.gz /beagle3/haky/users/temi/projects/TFXcan/data/chippeaks/weights/ENPACT_CHIPPEAKS_2026-01-02.compiled_weights.lambda.1se.txt.gz /scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.metadata.tsv /scratch/beagle3/temi/gene_modules/enpact_scores/GATA1_ACRs/GATA1_ACRs")
```


## Gather all predictions into 1

```{r}
calculate_regression_of_two_matrices <- function(ma, mb, matrix_names = c('matrixA', 'matrixB')){
    # match the names
    common_genes <- intersect(rownames(ma), rownames(mb))
    common_names <- intersect(colnames(ma), colnames(mb))

    MX <- ma[common_genes, common_names]
    MY <- mb[common_genes, common_names]

    if(all((nrow(MX) == nrow(MY)) & (ncol(MX) == ncol(MY)))){
        splitX <- base::asplit(MX, 1)
        splitY <- base::asplit(MY, 1)
        
        test_results <- mapply(function(x1, x2){
            tryCatch({
                lt <- data.table(expr = x1, binding = x2)
                test <- with(lt, lm(expr ~ binding)) %>% summary() %>% coef() %>% as.data.table(keep.rownames = 'term')
                return(test)
            }, error = function(err){
                return(NULL)
            })
        }, splitX, splitY, SIMPLIFY = FALSE)

        dt_results <- Filter(Negate(is.null), test_results) 
        
        return(dt_results)
    } else {
        stop("ERROR - Dimensions or names of input matrices are not the same")
    }
}
```

```{r}

```

```{r}
gexpr <- data.table::fread(file.path(tfxcan_directory, "experiments/gene_modules/files", "geuvadis.lcl.gene_expression.tsv.gz")) %>%
    dplyr::inner_join(dt.tss_metadata, by = c('gene' = 'locus')) %>%
    dplyr::relocate(external_gene_name, ensembl_gene_id) %>%
    dplyr::select(-c(external_gene_name, gene))
gexpr[1:5, 1:5]
```

```{r}
```

```{r}
GATA1_blood.enpact_scores <- data.table::fread("/scratch/beagle3/temi/gene_modules/enpact_scores/GATA1_ACRs/GATA1_ACRs.GATA1_Blood.enpact_scores.txt")
GATA1_blood.enpact_scores[1:5, 1:5]
```

```{r}
fastlm = function(xx,yy)
{
  ## compute betahat (regression coef) and pvalue with Ftest
  ## for now it does not take covariates

  df1 = 2
  df0 = 1
  ind = !is.na(xx) & !is.na(yy)
  xx = xx[ind]
  yy = yy[ind]
  n = sum(ind)
  xbar = mean(xx)
  ybar = mean(yy)
  xx = xx - xbar
  yy = yy - ybar

  SXX = sum( xx^2 )
  SYY = sum( yy^2 )
  SXY = sum( xx * yy )

  betahat = SXY / SXX

  RSS1 = sum( ( yy - xx * betahat )^2 )
  RSS0 = SYY

  fstat = ( ( RSS0 - RSS1 ) / ( df1 - df0 ) )  / ( RSS1 / ( n - df1 ) )
  pval = 1 - pf(fstat, df1 = ( df1 - df0 ), df2 = ( n - df1 ))
  res = list(betahat = betahat, pval = pval)

  return(res)
}
```

```{r}
exprMat <- gexpr %>% tibble::column_to_rownames('ensembl_gene_id') %>% as.matrix()
exprMat[1:5, 1:5]
```

```{r}
bindingMat <- GATA1_blood.enpact_scores %>% tibble::column_to_rownames('NAME') %>% as.matrix()
bindingMat[1:5, 1:5]
```

```{r}
cnames <- intersect(colnames(exprMat), colnames(bindingMat))
length(cnames)
```

```{r}
exprMat <- exprMat[, cnames]
bindingMat <- bindingMat[, cnames]

dim(exprMat); dim(bindingMat)
```


```{r}
exprMat %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "gene_expression.matrix.tsv"), row.names = T, col.names = T, quote = F, sep = '\t')

bindingMat %>% data.table::fwrite(file.path(tfxcan_directory, "experiments/gene_modules/files", "predicted_binding.matrix.tsv"), row.names = T, col.names = T, quote = F, sep = '\t')
```


```{r}
glue::glue("sbatch /beagle3/haky/users/temi/projects/TFXcan/src/correlate_binding_expression.acrs.sbatch /beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/predicted_binding.matrix.tsv /beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/gene_expression.matrix.tsv /beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/GATA1_ACRs.correlation")
```

```{r}
res.pvalues <- data.table::fread("/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/GATA1_ACRs.correlation.pvalues.tsv.gz")
dim(res.pvalues); res.pvalues[1:5, 1:5]
```

```{r}
dt.res.pvalues <- tidyr::pivot_longer(res.pvalues, cols = !V1, names_to = 'acr', values_to = 'pvalue') %>%
    dplyr::inner_join(dt.tss_metadata, by = c('V1' = 'ensembl_gene_id'))
dim(dt.res.pvalues); dt.res.pvalues[1:5, ]
```


```{r}
regulated_genes <- dt.res.pvalues %>%
    dplyr::filter(external_gene_name %in% GATA1.pert$gene_ID)

non_regulated_genes <- dt.res.pvalues %>%
    dplyr::filter(!external_gene_name %in% GATA1.pert$gene_ID)

dim(regulated_genes); dim(non_regulated_genes)
```

```{r}
pppvalues <- ggplot_qqplot(y = regulated_genes$pvalue, x = non_regulated_genes$pvalue) +
geom_abline(intercept = 0, slope = 1, col = 'red') +
theme_simple +
labs(y = 'ACRs target genes', x = 'ACRs non-target genes', subtitle = 'QQ plot of p values')

ggsave(file.path(figures_directory, '../qqplot.GATA1_ACRs.pvalues.png'), pppvalues, width = 3000, height = 3000, dpi = 800, units = 'px')
```

```{r}

```



```{r}
res.corrcoef <- data.table::fread("/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/GATA1_ACRs.correlation.corrcoef.tsv.gz")
dim(res.corrcoef); res.corrcoef[1:5, 1:5]
```

```{r}
dt.res.corrcoef <- tidyr::pivot_longer(res.corrcoef, cols = !V1, names_to = 'acr', values_to = 'corrcoef') %>%
    dplyr::inner_join(dt.tss_metadata, by = c('V1' = 'ensembl_gene_id'))
dim(dt.res.corrcoef); dt.res.corrcoef[1:5, ]
```

```{r}
regulated_genes <- dt.res.corrcoef %>%
    dplyr::filter(external_gene_name %in% GATA1.pert$gene_ID)

non_regulated_genes <- dt.res.corrcoef %>%
    dplyr::filter(!external_gene_name %in% GATA1.pert$gene_ID)

dim(regulated_genes); dim(non_regulated_genes)
```

```{r}
ggplot_qqplot(y = regulated_genes$corrcoef, x = non_regulated_genes$corrcoef) +
geom_abline(intercept = 0, slope = 1, col = 'red') +
theme_simple +
labs(y = 'ACRs target genes', x = 'ACRs non-target genes', subtitle = 'QQ plot of correlation coefficient')
```

```{r}
summary(abs(regulated_genes$corrcoef))
```

```{r}
summary(abs(non_regulated_genes$corrcoef))
```






```{r}
out <- apply(bindingMat[1:100, ], 1, function(col_b){
  apply(exprMat[1:100, ], 1, function(col2, col1) {
    cor.test(col2, col1)$p.value
  }, col1=col_b)
})
```

```{r}
min(out)
```

## Correlation with gene expression
```{r}

```
















```{r}
Xmat <- rhdf5::h5ls(file = "/scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.matrix.h5.gz")
h5closeAll()
```

```{r}
Xmat <- rhdf5::h5read(file = "/scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.matrix.h5.gz", name = '/matrix/block0_values')
Bmat <- rhdf5::h5read(file = "/scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.matrix.h5.gz", name = '/matrix/axis0')
```

```{r}
dim(Xmat); Xmat[1:10, 1:5]
```

```{r}
Mdata <- data.table::fread(file = "/scratch/beagle3/temi/gene_modules/epigenomes/GATA1_ACRs.epigenomes.processed.metadata.tsv")
```




Filter to genes in LCL

```{r}
avgenes <- TAL1.pert$gene_ID |> unique()
avgenes <- avgenes[which(avgenes %in% available.genes$external_gene_name)]
```


```{r}
dt_genes$log2FC |> summary()
```

```{r}
dt_genes.large <- dt_genes %>% dplyr::filter(abs(log2FC) > 1.5)
dim(dt_genes.large)
```

```{r}
dt_genes %>% dplyr::filter(perturbation_name %in% available.tfs$tf) %>% dplyr::filter(gene_ID %in% available.genes$external_gene_name)
```

```{r}
dt_accessibility <- data.table::fread(file.path(perturbation_directory, 'accessible_regions_tf.tsv')) 
```

```{r}
summary(dt_accessibility$log2FC)
```

```{r}
dt_accessibility.large <- dt_accessibility %>% dplyr::filter(abs(log2FC) > 1.5)
dim(dt_accessibility.large)
```

```{r}
dt_accessibility.large$perturbation_name %>% unique()
```













# this runs one locus at a time
# change to calculate LD
# https://merrimanlab.github.io/docs/locuszooms/multiple_lz_plots/




```{r}
# res.pvalues <- data.table::fread("/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/gene_expression.tf_vs_nontf_genes.pvalues.tsv.gz")
# dim(res.pvalues); res.pvalues[1:5, 1:5]

# res.corrcoef <- data.table::fread("/beagle3/haky/users/temi/projects/TFXcan/experiments/gene_modules/files/gene_expression.tf_vs_nontf_genes.corrcoef.tsv.gz")
# dim(res.corrcoef); res.corrcoef[1:5, 1:5]
```

```{r}
# dt.res.pvalues <- tidyr::pivot_longer(res.pvalues, cols = !V1, names_to = 'tf_gene', values_to = 'pvalue') %>%
#     dplyr::rename(gene = V1) %>%
#     dplyr::arrange(pvalue)
# dim(dt.res.pvalues); dt.res.pvalues[1:5, ]
```

```{r}
# dt.res.corrcoefs <- tidyr::pivot_longer(res.corrcoef, cols = !V1, names_to = 'tf_gene', values_to = 'corrcoef') %>%
#     dplyr::rename(gene = V1) %>%
#     dplyr::arrange(corrcoef)
# dim(dt.res.corrcoefs); dt.res.corrcoefs[1:5, ]
```

```{r}
# dt.res.merged <- dplyr::inner_join(dt.res.pvalues, dt.res.corrcoefs, by = c('tf_gene', 'gene')) %>%
#     dplyr::rename(target_gene = gene)
# dt.res.merged[1:5, ]; dim(dt.res.merged)


# ```{r}
# temp.merger <- dplyr::inner_join(dt.res.merged, dt.res.merged.2, by = c('tf_gene', 'target_gene'), suffix = c('.comp', '.incomp'))
# temp.merger[1:5, ]
# ```
# ```














- make heatmaps

```{r}
library(ComplexHeatmap)
library(cowplot)
```

```{r}
mat.corrcoef <- dt.res.merged.2 %>% pivot_wider(id_cols = tf_gene, names_from = target_gene, values_from = corr_coef) %>% tibble::column_to_rownames('tf_gene') %>% as.matrix()

dim(mat.corrcoef); mat.corrcoef[1:3, 1:3]
```

```{r}
matPlot <- mat.corrcoef[,]

# col_fun = circlize::colorRamp2(seq(quantile(matPlot, 0.01), quantile(matPlot, 0.99), len = 9), brewer.pal(9, "Reds"))
# brewer.pal(n = 8, name = "Reds")
```

```{r}
dim(matPlot)
```

```{r}
matOrder <- dt.res.merged.2 %>% dplyr::group_by(target_gene) %>%
    dplyr::summarize(medCorr = median(corr_coef, na.rm = TRUE)) %>%
    dplyr::arrange(desc(medCorr))
matOrder[1:5, ]; tail(matOrder)
```

```{r}
joinedMat <- cbind(matPlot[, matOrder$target_gene[1:300]], matPlot[, rev(rev(matOrder$target_gene)[1:300])])
dim(joinedMat)
```

```{r}
subset_pval_heatmap <- joinedMat %>% 
    ComplexHeatmap::Heatmap(col = col_fun, cluster_rows = F, cluster_columns = F, column_names_rot = 90, 
        heatmap_legend_param = list(title = expression("Corr")), row_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'), column_names_gp = grid::gpar(fontsize = 7, fontfamily = 'Helvetica'), rect_gp = grid::gpar(col = "black", lwd = 1),
        use_raster = TRUE, raster_resize_mat = TRUE, raster_by_magick = TRUE)
    # ComplexHeatmap::draw() %>% 
    # grid.grabExpr() %>%
    # ggdraw()

# subset_pval_heatmap
```

```{r}
# subset_pval_heatmap <- matPlot[1:3000, ] %>% densityHeatmap(show_quantiles = F, col = col_fun) 
# %>%
#     ComplexHeatmap::draw() %>% 
#     grid.grabExpr() %>%
#     ggdraw()
```

```{r}
png(file=file.path(figures_directory, "tfGene_vs_targetGene.heatmap.png"), width = 15000, height = 15000, units = 'px', res = 200)
ComplexHeatmap::draw(subset_pval_heatmap)
dev.off()
```



```{r}
# ggsave(file.path(figures_directory, "tfGene_vs_targetGene.heatmap.pdf"), subset_pval_heatmap, width = 48000, height = 100000, limitsize = FALSE)
```

- heatmap of correlation

```{r}
# corrHeatMapMat <- res.corrcoef %>%
#     tibble::column_to_rownames('V1') %>% as.matrix()

# uMat <- corrHeatMapMat[, 1:200]
# col_fun = circlize::colorRamp2(seq(quantile(uMat, 0.01), quantile(uMat, 0.99), len = 9), brewer.pal(9, "Reds"))

# corrHeatMapPlot <- uMat %>% 
#     ComplexHeatmap::Heatmap(col = col_fun, cluster_rows = T, cluster_columns = F, 
#         heatmap_legend_param = list(title = expression("Correlation\nCoefficient")), rect_gp = gpar(col = NA),
#         use_raster = TRUE, raster_resize_mat = TRUE, raster_by_magick = TRUE, show_column_names = FALSE, show_row_names = FALSE) %>%
#     ComplexHeatmap::draw() %>% 
#     grid.grabExpr() %>%
#     cowplot::ggdraw()

# corrHeatMapPlot

# ggsave(file.path(figures_directory, "tfGene_vs_targetGene.corr.heatmap.png"), corrHeatMapPlot, width = 48000, height = 48000, unit = 'px', limitsize = FALSE)
```

```{r}

```
